---
title: "Analyze voxel level growth"
author: "Ben Risk"
date: "2025-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reviewer comment

Currently, the analysis averages metrics across entire tracts. A more nuanced and informative analysis would involve examining spatially specific changes along the length of each tract, which could reveal regionally distinct developmental patterns.

```{r}
library(RNifti)
library(dplyr)
library(longCombat)

wm_mask = readNifti('../nifti/diffeo_warped/harmonized_dti_data/fa_mean_mask_thr015.nii.gz')
sum(wm_mask)
mean_fa = readNifti('../nifti/diffeo_warped/harmonized_dti_data/fa_mean.nii.gz')
sum(mean_fa>0.3)
# use more stringent threshold, or resample


# make 0 in mask equal to NA:
wm_mask_na=wm_mask
wm_mask_na[wm_mask==0]=NA

# get a template file with desired header info:
na_template = wm_mask
na_template[!is.na(wm_mask)]=NA

filepath='/home/ben/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/00008-03-02_CSI20151218_tensorup_s_m_aff_diffeo_masked_fa'

# get subject list:
dat = read.csv('dat_harmonized_lr.csv',stringsAsFactors = TRUE)
# restrict to GA 36 to 41:
dat = dat%>%filter(GA>=36 & GA<=41)

id_list = as.character(dat$Actual_ID)

# choose 1 axial slice:
axial_index = 90

# DTI measure:
measure.list = c('fa','tr','ad','rd')
measure=measure.list[1]

# create a matrix where each row represents a subject and each column represents a voxel:
nvoxel = sum(wm_mask[,,axial_index]==1)
dat_mat = matrix(NA,nrow=length(id_list),ncol=nvoxel)
```


# Loop through the 4 measures:

```{r}
for (measure in measure.list) {

  for (i in 1:length(id_list)) {  
    subj = readNifti(paste0('/home/ben/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/',id_list[i],'_tensorup_s_m_aff_diffeo_masked_',measure,'.nii.gz'))
    temp = subj[,,axial_index]
    dat_mat[i,] = temp[wm_mask[,,axial_index]==1]  
  }

  # Inspect data
  sum(dat_mat==0)
  # fa looks good, only six entries out of 400,000 equal to 0. don't worry about them.
  
  # convert trace to MD:
  if(measure=='tr') dat_mat = dat_mat/3
  
  # Merge voxel data with demographic data:
  dat_wide = data.frame('Actual_ID'=id_list,dat_mat)
  dat_wide$Actual_ID=factor(dat_wide$Actual_ID)
  dat_wide = merge(dat,dat_wide)
  
  # reshape wide to get input features for longitudinal combat
  feature_names=names(dat_wide)[244:length(dat_wide)]
  stopifnot(feature_names[1]=='X1')
  #detach('package:tidyr',unload=TRUE)
  degreesfreedom=5
  temp = splines::bs(dat_wide$Corr_age,df=degreesfreedom)
  temp = data.frame(temp)
  names(temp) = paste0("Corr_age",1:degreesfreedom)
  dat_wide = cbind(dat_wide,temp)
  
  # check the spline fit:
  # model.test = lm(dat_wide$AF_fa~as.matrix(temp))
  # plot(predict(model.test)~dat_wide$Corr_age)

  a = proc.time()     
  dat_combat <- longCombat(idvar='Unique_ID', 
                             timevar='Corr_age',
                             batchvar='Scanner', 
                             features=feature_names,
                             formula='Corr_age1 + Corr_age2+ Corr_age3 + Corr_age4 + Corr_age5+Sex', 
                            #formula='Corr_age + Sex', # note: linear vs spline seem similar
                             ranef='(1|Unique_ID)',
                             verbose=FALSE,
                             data=dat_wide)
  dat_harmonized <- merge(dat_wide, dat_combat$data_combat, by=c('Unique_ID', 'Corr_age','Scanner'))
  dat_harmonized$Family_ID=factor(dat_harmonized$Family_ID)
  save(dat_harmonized,file = paste0('../IntermediateData/',measure,'_harmonized.RData'))
}

```

```{r}
# Quick check things are okay:
p.mat.notharmonized=rep(NA,nvoxel)
p.mat.harmonized=rep(NA,nvoxel)
for (i in 1:nvoxel) {
  model.temp=summary(lm(dat_harmonized[,243+i]~dat_harmonized$Scanner))
  p.mat.notharmonized[i] = model.temp$coefficients[2,4]
  model.temp=summary(lm(dat_harmonized[,3476+i]~dat_harmonized$Scanner))
  p.mat.harmonized[i] = model.temp$coefficients[2,4]
}

sum(p.mat.notharmonized<0.05/nvoxel)
sum(p.mat.harmonized<0.05/nvoxel)
```
No p-values survive FWER correction in harmonized data, which is good.

# Fit GAMM and predict at 30, 90, and 180 and their derivatives:

```{r}
library(mgcv)
library(parallel)

corr_vals <- c(30, 90, 180)
h <- 1  # finite-difference step for Corr_age

newdata <- data.frame(
  Mean_RMS  = mean(dat_harmonized$Mean_RMS, na.rm = TRUE),
  Corr_age  = corr_vals,
  Unique_ID = factor(levels(dat_harmonized$Unique_ID)[1],
                     levels = levels(dat_harmonized$Unique_ID)),
  Family_ID = factor(levels(dat_harmonized$Family_ID)[1],
                     levels = levels(dat_harmonized$Family_ID))
)

idx <- 1:nvoxel
cores <- max(1, detectCores(logical = TRUE) - 1)

fit_one <- function(i,measure) {
  load(paste0('../IntermediateData/',measure,'_harmonized.RData'))
  df <- dat_harmonized
  df$y <- df[, 3476 + i]
  model.simple <- gam(
    y ~ s(Mean_RMS) + s(Corr_age) +
        s(Unique_ID, bs = "re") + s(Family_ID, bs = "re"),
    method = "REML",
    data = df
  )

  ## Population-level means (exclude RE)
  pred <- predict(
    model.simple,
    newdata = newdata,
    se.fit  = TRUE,
    type    = "response",
    exclude = c("s(Unique_ID)", "s(Family_ID)")
  )

  ## Derivatives wrt Corr_age at corr_vals (identity link ⇒ response == linear predictor)
  nd_plus  <- within(newdata, Corr_age <- Corr_age + h)
  nd_minus <- within(newdata, Corr_age <- Corr_age - h)

  Xp <- predict(model.simple, newdata = nd_plus,
                type = "lpmatrix", exclude = c("s(Unique_ID)", "s(Family_ID)"))
  Xm <- predict(model.simple, newdata = nd_minus,
                type = "lpmatrix", exclude = c("s(Unique_ID)", "s(Family_ID)"))
  D <- (Xp - Xm) / (2 * h)                 # central difference design

  beta <- coef(model.simple)
  Vb   <- vcov(model.simple)

  deriv <- as.numeric(D %*% beta)          # dμ/d(Corr_age)
  se_d  <- sqrt(rowSums((D %*% Vb) * D))   # sqrt(diag(D V D^T))

  list(
    mean        = as.numeric(pred$fit),
    lower       = as.numeric(pred$fit - 1.96 * pred$se.fit),
    upper       = as.numeric(pred$fit + 1.96 * pred$se.fit),
    deriv       = deriv,
    deriv_se    = se_d,
    deriv_lower = deriv - 1.96 * se_d,
    deriv_upper = deriv + 1.96 * se_d
  )
}

a <- proc.time()
for (measure in measure.list) {
    out_list <- mclapply(idx, fit_one, mc.cores = cores, measure=measure)
    timing <- proc.time() - a
    timing
    
    # Bind into matrices [nvoxel x 3]
    results_mean        <- do.call(rbind, lapply(out_list, `[[`, "mean"))
    results_lower       <- do.call(rbind, lapply(out_list, `[[`, "lower"))
    results_upper       <- do.call(rbind, lapply(out_list, `[[`, "upper"))
    results_deriv       <- do.call(rbind, lapply(out_list, `[[`, "deriv"))
    results_deriv_se    <- do.call(rbind, lapply(out_list, `[[`, "deriv_se"))
    results_deriv_lower <- do.call(rbind, lapply(out_list, `[[`, "deriv_lower"))
    results_deriv_upper <- do.call(rbind, lapply(out_list, `[[`, "deriv_upper"))
    
    # Label columns by Corr_age values
    labs <- paste0("Corr_", corr_vals)
    
    for (m in c("results_mean","results_lower","results_upper",
                "results_deriv","results_deriv_se",
                "results_deriv_lower","results_deriv_upper")) {
      tmp <- get(m)              # retrieve the object
      colnames(tmp) <- labs      # set the column names
      assign(m, tmp)             # put it back into the environment
    }
  save(results_mean,results_lower,results_upper,results_deriv,results_deriv_se,results_deriv_lower,results_deriv_upper,file=paste0('../IntermediateData/',measure,'_results_30_90_180.RData'))

# Build nifti files:

  arr4 <- array(NA_real_, dim = c(dim(na_template), 3))

  # repopulate a slice:
  for (t in 1:length(corr_vals)) {
    slice = matrix(NA,256,256)
    slice[wm_mask[,,90]==1]=results_mean[,t]  
    arr4[,,90,t]=slice
  }
  hdr = niftiHeader(na_template)
  spax = hdr$pixdim[2:4]
  TR = 1
  img4 <- asNifti(arr4, na_template,pixdim = c(NA, spax, TR))
  
  writeNifti(img4, file=paste0("~/Dropbox/WhiteMatterDevelopment/nifti/results/",measure,"_results.nii.gz"))
}  
proc.time()-a
```

Make derivatives niftis:

```{r}
for (measure in measure.list) {
  load(paste0("../IntermediateData/",measure,"_results_30_90_180.RData"))
  arr4 <- array(NA_real_, dim = c(dim(na_template), 3))

  # repopulate a slice:
  for (t in 1:length(corr_vals)) {
    slice = matrix(NA,256,256)
    slice[wm_mask[,,90]==1]=results_deriv[,t]  
    arr4[,,90,t]=slice
  }
  hdr = niftiHeader(na_template)
  spax = hdr$pixdim[2:4]
  TR = 1
  img4 <- asNifti(arr4, na_template,pixdim = c(NA, spax, TR))
  
  writeNifti(img4, file=paste0("~/Dropbox/WhiteMatterDevelopment/nifti/results/",measure,"_deriv.nii.gz"))
}
```

# Create exploratory plots

Examine how DTI metrics and their derivatives change across time:

```{r}
load('../IntermediateData/fa_results_30_90_180.RData')
fa_results_mean = results_mean
fa_results_deriv=results_deriv
colnames(fa_results_mean)
colnames(fa_results_deriv)
dim(fa_results_mean)

load('../IntermediateData/tr_results_30_90_180.RData')
md_results_mean = results_mean
md_results_deriv=results_deriv

load('../IntermediateData/rd_results_30_90_180.RData')
rd_results_mean = results_mean
rd_results_deriv=results_deriv

load('../IntermediateData/ad_results_30_90_180.RData')
ad_results_mean = results_mean
ad_results_deriv=results_deriv


hist(results_deriv[,1],xlim=c(-0.0006,0.0025))
hist(results_deriv[,2],xlim=c(-0.0006,0.0025))
hist(results_deriv[,3],xlim=c(-0.0006,0.0025))

hist(results_mean[,1])
hist(results_mean[,2])
hist(results_mean[,3])


plot(lowess(results_deriv[,1]~results_mean[,1]))
plot(lowess(results_deriv[,2]~results_mean[,2]))
plot(lowess(results_deriv[,3]~results_mean[,3]))


plot(results_deriv[,1]~results_mean[,1])
plot(results_deriv[,2]~results_mean[,2])
plot(results_deriv[,3]~results_mean[,3])

```


```{r}
library(tidyverse)
library(patchwork)

# ---- helper to pivot and label windows ----
to_long <- function(mat, metric, type) {
  as.data.frame(mat) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Window",
      values_to = "Value"
    ) %>%
    mutate(
      Metric = metric,
      Type   = type,
      Window = factor(
        Window,
        levels = c("Corr_30", "Corr_90", "Corr_180"),
        labels = c("30 Corr. age (days)", "90 Corr. age (days)", "180 Corr. age (days)")
      )
    )
}

# ---- plotting function (overlay densities) ----
density_plot <- function(df, title_text, xlab_text) {
  ggplot(df, aes(x = Value, color = Window, fill = Window)) +
  geom_density(
    alpha = 0.25,
    linewidth = 1,
    bw = "nrd0",
    adjust = 1.5
    ) +
  scale_fill_viridis_d(option = "plasma", end = 0.9) +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
    labs(
      title = title_text,
      x = xlab_text,
      y = "Density"
    ) +
    theme_classic(base_size = 14) +
    theme(
      legend.title = element_blank()
    )
}

# ---- build one long dataframe (optional, but handy) ----
df_all <- bind_rows(
  to_long(fa_results_mean,  "FA", "Mean"),
  to_long(fa_results_deriv, "FA", "Derivative"),
  to_long(md_results_mean,  "MD", "Mean"),
  to_long(md_results_deriv, "MD", "Derivative"),
  to_long(rd_results_mean,  "RD", "Mean"),
  to_long(rd_results_deriv, "RD", "Derivative"),
  to_long(ad_results_mean,  "AD", "Mean"),
  to_long(ad_results_deriv, "AD", "Derivative")
)

# ---- create the 8 plots (2 rows x 4 columns) ----
p_fa_mean  <- density_plot(filter(df_all, Metric=="FA", Type=="Mean"),       "FA", "Estimated FA value")
p_md_mean  <- density_plot(filter(df_all, Metric=="MD", Type=="Mean"),       "MD", "Estimated MD value")
p_rd_mean  <- density_plot(filter(df_all, Metric=="RD", Type=="Mean"),       "RD", "Estimated RD value")
p_ad_mean  <- density_plot(filter(df_all, Metric=="AD", Type=="Mean"),       "AD", "Estimated AD value")

p_fa_deriv <- density_plot(filter(df_all, Metric=="FA", Type=="Derivative"), "FA", "Estimated derivative")
p_md_deriv <- density_plot(filter(df_all, Metric=="MD", Type=="Derivative"), "MD", "Estimated derivative")
p_rd_deriv <- density_plot(filter(df_all, Metric=="RD", Type=="Derivative"), "RD", "Estimated derivative")
p_ad_deriv <- density_plot(filter(df_all, Metric=="AD", Type=="Derivative"), "AD", "Estimated derivative")

# ---- assemble into 2x4 with one shared legend ----
panel_2x4 <-
  (p_fa_mean | p_md_mean | p_ad_mean | p_rd_mean) /
  (p_fa_deriv | p_md_deriv | p_ad_deriv | p_rd_deriv) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_2x4

ggsave("../Figures/dti_density_2x4.png", panel_2x4, width = 16, height = 8, dpi = 300)
```