---
title: "White Matter Development Data Preparation and Harmonization"
author: "Benjamin Risk"
# date: "2/14/2022"
date: 07/31/2025
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Updates:
-Added Family ID

-Added Actual_ID which is used in merge to motion data

-Added mean absolute and relative motion

-Reduced to demographic and motion variables

## Data prep

Grab the IDs, DOB, GA, chor_age, Corr_age, and Sex

```{r,results=FALSE}
library(dplyr)
dat = read.delim('../Data/dti_data_extracted_by_tracts_all_atlas_symmetry_xyflipped_2_TnP2020.txt',sep=" ",stringsAsFactors = TRUE)
dat$Scanner = as.factor(ifelse(dat$ifPrisma==0,'trio','prisma'))

```



There are issues with Corr_age for one infant. Replace with correct corrected age.

```{r}
# evidence that correction is needed:
dat_short = dat[,c('Unique_ID','chor_age','Corr_age','GA')]
dat_short$my_correctedage = dat$chor_age-(40-dat$GA)*7
dat_short$diff = dat_short$Corr_age - dat_short$my_correctedage

dat_short[which(abs(dat_short$diff)>1),]
# note: GA=37.3 for one infant is correct

# overwrite the old Corr_age with correct Corr_age:
dat$Corr_age = dat$chor_age-(40-dat$GA)*7
```

Subset to columns used in downstream analyses:

```{r}
dat = dat[,c('Actual_ID','DOB','GA','chor_age','Corr_age','Sex','Unique_ID','ifPrisma','Scanner')]
```

## Added 7/30/2025:

Create family ID variable:

```{r}
# how many pairs of twins due we have:
dat$Family_ID = factor(substr(dat$Unique_ID,1,5))

dat = dat[!duplicated(dat),]
family_counts <- dat %>%
  distinct(Unique_ID, .keep_all = TRUE) %>%       # Keep one row per Unique_ID
  count(Family_ID, name = "n_in_family")       # Count number of Unique_IDs per FamilyID

dat <- dat %>%
  left_join(family_counts, by = "Family_ID")%>%
  mutate(Sibling = as.integer(n_in_family > 1))

 
table(dat$n_in_family)
# we have 30 scans from siblings (with some twins)
```

## Added 7/31/25:

Calculate Mean Absolute and Relative Motion:

```{r}
motion_data = dat[,c('Actual_ID','Unique_ID')]
motion_data$Absolute_RMS=motion_data$Mean_RMS=NA

for (i in 1:nrow(motion_data)) {
  rms=read.table(paste0('/home/ben/Dropbox/WhiteMatterDevelopment/EddyMotionParameters/',motion_data$Actual_ID[i],'/eddy_corrected_dataup.eddy_movement_rms'))
  motion_data[i,'Absolute_RMS'] = mean(rms[2:nrow(rms),1])
  motion_data[i,'Mean_RMS'] = mean(rms[2:nrow(rms),2])
}
motion_data = motion_data%>%select(-Unique_ID)
dat=merge(dat,motion_data)
```

Write file to disk:
```{r}
write.csv(dat,file='dat_demog.csv',row.names=FALSE)
```