---
title: "White Matter Development Data Preparation and Harmonization"
author: "Benjamin Risk"
# date: "2/14/2022"
date: 12/16/2025
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data prep



```{r,results=FALSE}
library(RNifti)
library(readxl)
library(dplyr)
library(stringr)

a = proc.time()
atlas = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/Register_Tracts_from_Template_to_Current_Space/Template_and_Tracts_Original_Asymmetry/tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz')
image(atlas[,,69])
```

Update this atlas with the updated arcuate fasciculus that corrects the R-L issues
```{r}
```

overwrite tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz with new AF:

```{r}
af_liberal=TRUE

if(af_liberal) {
  af = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/AF_liberal/AF_L_R_liberal_flipped_Warp.nii.gz')
  
  # Based on visual inspection of AF_L_R_liberal_flipped_Warp.nii.gz, "1" corresponds to AF right, and "2" to AF left. Correct the labels to correspond to tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz:
  
  table(af)
  af[af==2]=9
  af[af==1]=2
  af[af==9]=1
  atlas_old=atlas
  atlas[atlas_old==1 | atlas_old==2] = 0
  atlas[af==1] = af[af==1]
  atlas[af==2] = af[af==2]
}
table(atlas)

```

read in example participant:
```{r}
sub_fa = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/00008-03-02_CSI20151218_tensorup_s_m_aff_diffeo_masked_fa.nii.gz')
```
ben's notes: I checked, and the files  00008-03-02_CSI20151218_tensorup_s_m_aff_diffeo_masked_fa.nii represents the not harmonized data. 

read in labels and "symmetric" data used in main analysis for auditing:
```{r}
labels = read_xlsx("~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/Table_Included_tracts reorderedCol.xlsx")
labels = labels%>%rename(ROI_number = `Intensity Index (Asymmetry)`,ROI_name=`Abbreviations`)
# sort by ROI_number:
labels = labels[order(labels$ROI_number),]

dat_demog = read.csv('dat_demog.csv')
actual_ids = dat_demog$Actual_ID
metrics = c('fa','tr','ad','rd')
var_names = labels$ROI_name
roi_labels = labels$ROI_number
```

Compute asymmetric data:
  
```{r}  
# A function that returns the file path for a given subject ID and metric.
file_template <- "~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/%s_tensorup_s_m_aff_diffeo_masked_%s.nii.gz"                                
# Example expected paths like "/data/dti/ACT123/fa.nii.gz"
file_for <- function(id, metric) sprintf(file_template, id, metric)
# making nice progress here, working through chat gpt code

stopifnot(length(var_names) == length(roi_labels))
stopifnot(is.array(atlas))

# Precompute voxel indices for each ROI (fast later)
roi_index_list <- lapply(roi_labels, function(lbl) which(atlas == lbl))
names(roi_index_list) <- var_names


# Helper safe mean
.safe_mean <- function(x) if (length(x)) mean(x, na.rm = TRUE) else NA_real_
# ben note: see if we can add a tolerance here


# ---- Prepare output frame: Actual_ID + 19*4 = 76 metric columns ----
all_metric_cols <- as.vector(outer(var_names, metrics, function(n, m) paste0(n, "_", m)))
out <- data.frame(Actual_ID = actual_ids, matrix(NA_real_, nrow = length(actual_ids), ncol = length(all_metric_cols)))
colnames(out)[-1] <- all_metric_cols

# ---- Core loop: actual_ids × metrics ----
for (i in seq_along(actual_ids)) {
  id <- actual_ids[i]

  # Optional: message progress
  if (i %% 10 == 0) message("Scan ", i, "/", length(actual_ids), ": ", id)

  for (met in metrics) {
    img_path <- file_for(id, met)
    if (!file.exists(img_path)) {
      warning("Missing file: ", img_path, " (leaving NAs for this subject/metric)")
      next
    }

    img <- readNifti(img_path)

    # Dimension check (basic guard)
    if (!all(dim(img) == dim(atlas))) {
      stop("Dimension mismatch for subject ", id, ", metric ", met,
           ": img dims ", paste(dim(img), collapse = "x"),
           " vs atlas dims ", paste(dim(atlas), collapse = "x"))
    }

    # Compute means per ROI using precomputed indices
    roi_means <- vapply(roi_index_list, function(idx) .safe_mean(img[idx]), numeric(1))
    # to remove 0s: .safe_mean(img[idx][img[idx] != 0])
    
    # Place into correct columns
    cols_this_metric <- paste0(names(roi_means), "_", met)
    out[i, cols_this_metric] <- roi_means
    }
}

out <- out %>%
  rename_with(.fn = ~ sub("_tr$", "_md", .x), .cols = ends_with("_tr")) %>%
  mutate(across(ends_with("_md"), ~ .x / 3))

proc.time()-a
```


## Calculate the symmetric (non-lateralized) version for AF, ATR, Ci, Fx, IFOF, ILF, PT, and UF:

```{r}
# ---- voxel counts for each ROI (fixed across subjects) ----
roi_nvox <- vapply(roi_index_list, length, integer(1))  # names match var_names
metrics=c('fa','md','ad','rd')

# ---- identify tracts with L/R versions based on var_names ----
bases_L <- str_match(var_names, "^(.*)_L$")[,2]
bases_R <- str_match(var_names, "^(.*)_R$")[,2]
lr_bases <- intersect(na.omit(bases_L), na.omit(bases_R))

# Safety: show what will be combined
message("Combining L/R for: ", paste(lr_bases, collapse = ", "))

# ---- add weighted-mean combined columns for each metric ----
for (b in lr_bases) {
  nL <- roi_nvox[paste0(b, "_L")]
  nR <- roi_nvox[paste0(b, "_R")]
  denom <- nL + nR

  for (met in metrics) {
    colL <- paste0(b, "_L_", met)
    colR <- paste0(b, "_R_", met)
    colC <- paste0(b, "_",   met)
    out[[colC]] <- (out[[colL]] * nL + out[[colR]] * nR) / denom
  }
}
```

Calculate whole brain DTI:
```{r}
# ---- Whole-brain weighted means (STRICT: any NA => NA) ----

for (met in metrics) {
  roi_cols <- paste0(var_names, "_", met)   # e.g., AF_L_fa, ..., UF_R_fa
  w <- roi_nvox[var_names]                  # voxel counts aligned to var_names
  W <- sum(w)

  X <- as.matrix(out[, roi_cols, drop = FALSE])

  # If any ROI is NA in a row, WB is NA for that subject
  out[[paste0("WB_", met)]] <-
    as.numeric(X %*% w) / W

  
  stopifnot(all(colnames(out)[match(paste0(var_names, "_", metrics[1]),
                                  colnames(out))] %in% colnames(out)))
  stopifnot(length(w) == ncol(X))
}
```

## Apply longitudinal combat and investigate scanner effects

```{r}
dat_wide = merge(dat_demog,out,by="Actual_ID", suffixes = c("", ".y"))
```


Construct a b-spline for corrected age. Then longitudinal combat will use this spline for corrected age. We used df=5 to capture a relatively simple curve, since the effective degrees of freedom from initial GAMMs indicated effective degrees of freedom of 3-4 in the penalized spline approach. Based on inspection, the results appear relatively robust to this choice, although the spline may help to reduce some scanner effects. 
Note, longitudinal combat models the batch effect on the mean as an intercept, i.e., shift up or down in the developmental trajectory. Current harmonization methods do not allow batch effects to cause different curves. Thus, it is important to check that there are not substantial interactions between batch (here, scanner) and the develomental trajectory, i.e., that the curves do not substantially differ on the trio and prisma scanners.

Note to self: watch out for the use of indexing on column names. Corpus collosum 
```{r}
library(longCombat)
# grab variables names:
var_names_metrics = names(dat_wide)[15:126]
# check names
stopifnot(var_names_metrics[1]=='AF_L_fa')
stopifnot(var_names_metrics[length(var_names_metrics)]=='WB_rd')

# detach('package:tidyr',unload=TRUE)

degreesfreedom=5
temp = splines::bs(dat_wide$Corr_age,df=degreesfreedom)
temp = data.frame(temp)
names(temp) = paste0("Corr_age",1:degreesfreedom)
dat_wide = cbind(dat_wide,temp)


dat_combat <- longCombat(idvar='Unique_ID', 
                             timevar='Corr_age',
                             batchvar='Scanner', 
                             features=var_names_metrics,
                             formula='Corr_age1 + Corr_age2+ Corr_age3 + Corr_age4 + Corr_age5+Sex', 
                            #formula='Corr_age + Sex', # note: linear vs spline seem similar
                             ranef='(1|Unique_ID)',
                             verbose=FALSE,
                             data=dat_wide)
dat_harmonized <- merge(dat_wide, dat_combat$data_combat, by=c('Unique_ID','Corr_age','Scanner'))
``` 

Write file to disk:
```{r}
write.csv(dat_harmonized,file='dat_harmonized_lr.csv',row.names=FALSE)
```
