---
title: 'White Matter Development: Analysis'
author: "Benjamin Risk"
date: "01/15/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-correct titles on tract specific bands
-Create new version of figure 5, separate from sex effects


```{r}
library(mgcv)
library(itsadug)
dat_harmonized = read.csv('dat_harmonized.csv',stringsAsFactors = TRUE)
```

### FA

This function extracts the largest age at which there is a significant difference:
```{r}
# note: full term is 37
maxChronAge = function(GAlow,GAhigh=40,model,ylab,less=TRUE) {
  a=plot_diff(model,view='chor_age',comp=list(GA=c(GAlow,GAhigh)),sim.ci = TRUE, hide.label=TRUE,xlab='Chronological age (days)',ylab=ylab)
  if(less) {
  max(a$chor_age[a$est+a$sim.CI<0])
  } else {
    max(a$chor_age[a$est-a$sim.CI>0])
  }
}
```

The syntax used here is described as the functional anova in the supporting R documentation for ?ti(). 
```{r,error=TRUE}
model0.fa = gam(WB_fa.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)

    
# print out the k's:
gam.check(model0.fa)
# edf are all far from k', so the default k's suffice. 
summary(model0.fa)

  # alternative syntax from gam book, not the same as the syntax used in documentation:
  model0.fa.check = gam(WB_fa.combat~s(chor_age,k=9)+s(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
  summary(model0.fa.check)
  # results are not identical, but very similar. Use the ti(x) + ti(z) + ti(x,z) described in ?ti()
  
  
  # check it is location invariant:
  dat_harmonized$c_chor_age = scale(dat_harmonized$chor_age,center=TRUE,scale=FALSE)
  dat_harmonized$c_GA = scale(dat_harmonized$GA,center=TRUE,scale=FALSE)
  
  model0.fa.audit = gam(WB_fa.combat~ti(c_chor_age,k=9)+ti(c_GA,k=5)+ti(c_chor_age,c_GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
  summary(model0.fa.audit)
  # good. 



vis.gam(model0.fa,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='FA')

set.seed(123)

tiff('../Figures/DiffCurves_FA.tiff',width=720,height=480,pointsize=20)
par(mfrow=c(2,3))
(ga.35.fa=maxChronAge(GAlow=35,GAhigh=40,model=model0.fa,ylab='Est. difference in FA'))
(ga.36.fa=maxChronAge(GAlow=36,GAhigh=40,model=model0.fa,ylab='Est. difference in FA'))
(ga.37.fa=maxChronAge(GAlow=37,GAhigh=40,model=model0.fa,ylab='Est. difference in FA'))
(ga.38.fa=maxChronAge(GAlow=38,GAhigh=40,model=model0.fa,ylab='Est. difference in FA'))
(ga.39.fa=maxChronAge(GAlow=39,GAhigh=40,model=model0.fa,ylab='Est. difference in FA'))
(ga.41.fa=maxChronAge(GAlow=40,GAhigh=41,model=model0.fa,ylab='Est. difference in FA'))
dev.off()

```

### MD
```{r,error=TRUE}
model0.md = gam(WB_md.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)

gam.check(model0.md) # edf are all far from k' 
summary(model0.md)


tiff('../Figures/DiffCurves_MD.tiff',width=720,height=480,pointsize=20)
par(mfrow=c(2,3))
(ga.35.md=maxChronAge(GAlow=35,GAhigh=40,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
(ga.36.md=maxChronAge(GAlow=36,GAhigh=40,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
(ga.37.md=maxChronAge(GAlow=37,GAhigh=40,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
(ga.38.md=maxChronAge(GAlow=38,GAhigh=40,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
(ga.39.md=maxChronAge(GAlow=39,GAhigh=40,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
(ga.41.md=maxChronAge(GAlow=40,GAhigh=41,model=model0.md,less=FALSE,ylab='Est. difference in MD'))
dev.off()


#par(mfrow=c(1,2))
#vis.gam(model0.md,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='Trace')
```

There is not a significant interaction between chronological age and gestational age for md (p=0.09), and the lines look more parallel than in FA. Nonetheless, there is some evidence of an interaction, as the lines becomes somewhat closer together near 200.

### AD
```{r,error=TRUE}
model0.ad = gam(WB_ad.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.ad) # edf are all far from k', so the default k's suffice. 
summary(model0.ad)

plot_diff(model0.ad,view='chor_age',comp=list(GA=c(36,40)),sim.ci = TRUE)

tiff('../Figures/DiffCurves_AD.tiff',width=720,height=480,pointsize=20)
par(mfrow=c(2,3))
(ga.35.ad=maxChronAge(GAlow=35,GAhigh=40,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
(ga.36.ad=maxChronAge(GAlow=36,GAhigh=40,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
(ga.37.ad=maxChronAge(GAlow=37,GAhigh=40,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
(ga.38.ad=maxChronAge(GAlow=38,GAhigh=40,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
(ga.39.ad=maxChronAge(GAlow=39,GAhigh=40,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
(ga.41.ad=maxChronAge(GAlow=40,GAhigh=41,model=model0.ad,less=FALSE,ylab='Est. difference in AD'))
dev.off()

vis.gam(model0.ad,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='AD')

```

### RD
```{r,error=TRUE}
model0.rd = gam(WB_rd.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.rd) # edf are all far from k', so the default k's suffice. 
summary(model0.rd)

vis.gam(model0.rd,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='RD')

tiff('../Figures/DiffCurves_RD.tiff',width=720,height=480,pointsize=20)
par(mfrow=c(2,3))
(ga.35.rd=maxChronAge(GAlow=35,GAhigh=40,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
(ga.36.rd=maxChronAge(GAlow=36,GAhigh=40,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
(ga.37.rd=maxChronAge(GAlow=37,GAhigh=40,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
(ga.38.rd=maxChronAge(GAlow=38,GAhigh=40,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
(ga.39.rd=maxChronAge(GAlow=39,GAhigh=40,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
(ga.41.rd=maxChronAge(GAlow=40,GAhigh=41,model=model0.rd,less=FALSE,ylab='Est. difference in RD'))
dev.off()

```
There is a trend interaction between chronological and gestation age on RD, but overall the lines are parallel. 

### Plots of gestational age at chronological age = 30:
```{r}

tiff('../Figures/GestationalAge_30_90_180.tiff',width=1080,height=1280,pointsize=18)
par(mfrow=c(4,4))
focal_chron=30
plot_smooth(model0.fa, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain FA  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)'
, hide.label=TRUE,main='A',xlim=c(35,41),rug=FALSE,col='slateblue1')

plot_smooth(model0.md, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain MD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='B',xlim=c(35,41),rug=FALSE,col='slateblue1')

plot_smooth(model0.ad, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain AD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='C',xlim=c(35,41),rug=FALSE,col='slateblue1')

plot_smooth(model0.rd, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain RD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='D',xlim=c(35,41),rug=FALSE,col='slateblue1')

focal_chron=90
plot_smooth(model0.fa, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain FA  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)', hide.label=TRUE,main='E',xlim=c(35,41),rug=FALSE,col='slateblue3')

plot_smooth(model0.md, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain MD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='F',xlim=c(35,41),rug=FALSE,col='slateblue3')

plot_smooth(model0.ad, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain AD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='G',xlim=c(35,41),rug=FALSE,col='slateblue3')

plot_smooth(model0.rd, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain RD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='H',xlim=c(35,41),rug=FALSE,col='slateblue3')

focal_chron=180
plot_smooth(model0.fa, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain FA  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)', hide.label=TRUE,main='I',xlim=c(35,41),rug=FALSE,col='darkslateblue')

plot_smooth(model0.md, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain MD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='J',xlim=c(35,41),rug=FALSE,col='darkslateblue')

plot_smooth(model0.ad, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain AD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='K',xlim=c(35,41),rug=FALSE,col='darkslateblue')

plot_smooth(model0.rd, view="GA",cond=list(chor_age=focal_chron),lwd=2,sim.ci=TRUE,ylab=paste0('Whole-brain RD  age=',focal_chron,' days'),xlab='Gestational age at birth (weeks)',hide.label=TRUE,main='L',xlim=c(35,41),rug=FALSE,col='darkslateblue')

### Plots of growth curves for the four dti measures at different GA:

#tiff('../Figures/GestandChronAgeFAMDADRD.tiff',width=1080,height=320,pointsize=18)
#par(mfrow=c(1,4))

######### FA:
cColors = viridis::plasma(15)
plot_smooth(model0.fa, view="chor_age",cond=list(GA=35),col=cColors[1],rug=FALSE,ylim=c(0.12,0.205),lwd=2,se=0,ylab='Whole-brain FA',xlab='Chronological age (days)',hide.label=TRUE,main='M',xlim=c(10,202))

plot_smooth(model0.fa, view="chor_age",cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2,se=0)

plot_smooth(model0.fa, view="chor_age",  cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2,se=0)

plot_smooth(model0.fa, view="chor_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.fa, view="chor_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.fa, view="chor_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.fa, view="chor_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)

legend(x='bottomright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1)

abline(v=ga.35.fa,col=cColors[1],lwd=2,lty=2)
abline(v=ga.36.fa,col=cColors[3],lwd=2,lty=2)
abline(v=ga.37.fa,col=cColors[5],lwd=2,lty=2)
abline(v=ga.38.fa,col=cColors[7],lwd=2,lty=2)
abline(v=ga.39.fa,col=cColors[9],lwd=2,lty=2)
abline(v=ga.41.fa,col=cColors[13],lwd=2,lty=2)


######### MD:
plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=35),ylab='Whole-brain MD',xlab='Chronological age (days)',hide.label=TRUE,col=cColors[1],rug=FALSE,main='N',lwd=2,ylim=c(1,1.5))

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.md, view="chor_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)

abline(v=ga.35.md,col=cColors[1],lwd=2,lty=2)
abline(v=ga.36.md,col=cColors[3],lwd=2,lty=2)
abline(v=ga.37.md,col=cColors[5],lwd=2,lty=2)
abline(v=ga.38.md,col=cColors[7],lwd=2,lty=2)
abline(v=ga.39.md,col=cColors[9],lwd=2,lty=2)
abline(v=ga.41.md,col=cColors[13],lwd=2,lty=2)


######### AD:
plot_smooth(model0.ad, view="chor_age", se=0,ylab='Whole-brain AD',xlab='Chronological age (days)',hide.label=TRUE, cond=list(GA=35),col=cColors[1],rug=FALSE,main='O',lwd=2,ylim=c(1.2,1.65))

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.ad, view="chor_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)
abline(v=ga.35.ad,col=cColors[1],lwd=2,lty=2)
abline(v=ga.36.ad,col=cColors[3],lwd=2,lty=2)
abline(v=ga.37.ad,col=cColors[5],lwd=2,lty=2)
abline(v=ga.38.ad,col=cColors[7],lwd=2,lty=2)
abline(v=ga.39.ad,col=cColors[9],lwd=2,lty=2)
abline(v=ga.41.ad,col=cColors[13],lwd=2,lty=2)



######### RD:
plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=35),hide.label=TRUE,xlab='Chronological age (days)',ylab='Whole-brain RD',col=cColors[1],add=FALSE,rug=FALSE, ylim=c(0.9,1.35),main='P',lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.rd, view="chor_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)
#b+legend(x='topright',legend=c('GA=32','GA=34','GA=36','GA=38','GA=40'),lty=1,col=cColors[seq(1,11,by=2)],bty='n',cex=0.5)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)

abline(v=ga.35.rd,col=cColors[1],lwd=2,lty=2)
abline(v=ga.36.rd,col=cColors[3],lwd=2,lty=2)
abline(v=ga.37.rd,col=cColors[5],lwd=2,lty=2)
abline(v=ga.38.rd,col=cColors[7],lwd=2,lty=2)
abline(v=ga.39.rd,col=cColors[9],lwd=2,lty=2)
abline(v=ga.41.rd,col=cColors[13],lwd=2,lty=2)


dev.off()
```

## Interaction between GA and corrected age
Next, we check whether there is a significant interaction between gestational age and corrected age in FA. 
```{r}
model0.corr.fa = gam(WB_fa.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.corr.fa)
summary(model0.corr.fa)

model0.corr.md = gam(WB_md.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.corr.md)
summary(model0.corr.md)

model0.corr.ad = gam(WB_ad.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.corr.ad)
summary(model0.corr.ad)


model0.corr.rd = gam(WB_rd.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
gam.check(model0.corr.rd)
summary(model0.corr.rd)
```

Main effect of gestational age is no longer significant. However, some interaction effects now appear in the interaction term with gestational age. Create plots to examine. 

```{r}
tiff('../Figures/GestandCorrectedAgeFAMDADRD.tiff',width=1080,height=320,pointsize=18)
par(mfrow=c(1,4))

######### FA:
cColors = viridis::plasma(15)
plot_smooth(model0.corr.fa, view="Corr_age",cond=list(GA=35),col=cColors[1],rug=FALSE,ylim=c(0.12,0.205),lwd=2,se=0,ylab='Whole-brain FA',xlab='Corrected age (days)',hide.label=TRUE,main='A',xlim=c(10,202))

plot_smooth(model0.corr.fa, view="Corr_age",cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2,se=0)

plot_smooth(model0.corr.fa, view="Corr_age",  cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2,se=0)

plot_smooth(model0.corr.fa, view="Corr_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.fa, view="Corr_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.fa, view="Corr_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.fa, view="Corr_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)

legend(x='bottomright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1)

# abline(v=ga.35.fa,col=cColors[1],lwd=2,lty=2)
# abline(v=ga.36.fa,col=cColors[3],lwd=2,lty=2)
# abline(v=ga.37.fa,col=cColors[5],lwd=2,lty=2)
# abline(v=ga.38.fa,col=cColors[7],lwd=2,lty=2)
# abline(v=ga.39.fa,col=cColors[9],lwd=2,lty=2)
# abline(v=ga.41.fa,col=cColors[13],lwd=2,lty=2)


######### MD:
plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=35),ylab='Whole-brain MD',xlab='Corrected age (days)',hide.label=TRUE,col=cColors[1],rug=FALSE,main='B',lwd=2,ylim=c(1,1.5))

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.md, view="Corr_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)

# abline(v=ga.35.md,col=cColors[1],lwd=2,lty=2)
# abline(v=ga.36.md,col=cColors[3],lwd=2,lty=2)
# abline(v=ga.37.md,col=cColors[5],lwd=2,lty=2)
# abline(v=ga.38.md,col=cColors[7],lwd=2,lty=2)
# abline(v=ga.39.md,col=cColors[9],lwd=2,lty=2)
# abline(v=ga.41.md,col=cColors[13],lwd=2,lty=2)

######### AD:
plot_smooth(model0.corr.ad, view="Corr_age", se=0,ylab='Whole-brain AD',xlab='Corrected age (days)',hide.label=TRUE, cond=list(GA=35),col=cColors[1],rug=FALSE,main='C',lwd=2,ylim=c(1.2,1.65))

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.ad, view="Corr_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)
# abline(v=ga.35.ad,col=cColors[1],lwd=2,lty=2)
# abline(v=ga.36.ad,col=cColors[3],lwd=2,lty=2)
# abline(v=ga.37.ad,col=cColors[5],lwd=2,lty=2)
# abline(v=ga.38.ad,col=cColors[7],lwd=2,lty=2)
# abline(v=ga.39.ad,col=cColors[9],lwd=2,lty=2)
# abline(v=ga.41.ad,col=cColors[13],lwd=2,lty=2)



######### RD:
plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=35),hide.label=TRUE,xlab='Corrected age (days)',ylab='Whole-brain RD',col=cColors[1],add=FALSE,rug=FALSE, ylim=c(0.9,1.35),main='D',lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2)

plot_smooth(model0.corr.rd, view="Corr_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2)

#legend(x='topright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,col=cColors[seq(1,13,by=2)],bty='n',cex=1,lwd=2)

# abline(v=ga.35.rd,col=cColors[1],lwd=2,lty=2)
# abline(v=ga.36.rd,col=cColors[3],lwd=2,lty=2)
# abline(v=ga.37.rd,col=cColors[5],lwd=2,lty=2)
# abline(v=ga.38.rd,col=cColors[7],lwd=2,lty=2)
# abline(v=ga.39.rd,col=cColors[9],lwd=2,lty=2)
# abline(v=ga.41.rd,col=cColors[13],lwd=2,lty=2)

dev.off()
```


## Sex effects 

### Sex models
```{r}

model.sex.fa = gam(WB_fa.combat~s(Corr_age,by=Sex)+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
summary(model.sex.fa)

model.sex.md = gam(WB_md.combat~s(Corr_age,by=Sex)+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
summary(model.sex.md)

model.sex.ad = gam(WB_ad.combat~s(Corr_age)+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
summary(model.sex.ad)

model.sex.rd = gam(WB_rd.combat~s(Corr_age,by=Sex)+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
summary(model.sex.rd)
```

### Examine differences
```{r}
plot_diff(model.sex.fa,view='Corr_age',comp=list(Sex=c('F','M')),ylab='Difference in FA',xlab='Corrected age (days)',hide.label=TRUE,sim.ci=TRUE)

plot_diff(model.sex.md,view='Corr_age',comp=list(Sex=c('F','M')),ylab='Difference in MD',xlab='Corrected age (days)',hide.label=TRUE,sim.ci=TRUE)

plot_diff(model.sex.ad,view='Corr_age',comp=list(Sex=c('F','M')),ylab='Difference in AD',xlab='Corrected age (days)',hide.label=TRUE,sim.ci=TRUE)

plot_diff(model.sex.rd,view='Corr_age',comp=list(Sex=c('F','M')),ylab='Difference in RD',xlab='Corrected age (days)',hide.label=TRUE,sim.ci=TRUE)
```
Sex differences from 2 to 27 weeks are significant in FA.

Sex differences are significant in AD.

### Plots
```{r}
tiff('../Figures/CorrectedAgeBySexWB.tiff',width=1080,height=320,pointsize=18)
par(mfrow=c(1,4))
plot_smooth(model.sex.fa, view="Corr_age", cond=list(Sex='M'), col='blue',print.summary=FALSE,hide.label=TRUE,ylab='Whole-brain FA',xlab='Corrected age (days)',sim.ci=TRUE,main='A')
plot_smooth(model.sex.fa,view="Corr_age",cond=list(Sex='F'),col='red',print.summary=FALSE,add=TRUE,sim.ci=TRUE)
legend(x='bottomright',legend=c('Males','Females'),lty=1,col=c('blue','red'),bty='n',cex=1,lwd=2)
abline(v=27,lty=2,col='red',lwd=2)

plot_smooth(model.sex.md, view="Corr_age", cond=list(Sex='M'), col='blue',print.summary=FALSE,hide.label=TRUE,ylab='Whole-brain MD',xlab='Corrected age (days)',sim.ci=TRUE,main='B')
plot_smooth(model.sex.md,view="Corr_age",cond=list(Sex='F'),col='red',print.summary=FALSE,add=TRUE,sim.ci=TRUE)

plot_smooth(model.sex.ad, view="Corr_age", cond=list(Sex='M'), col='blue',print.summary=FALSE,hide.label=TRUE,ylab='Whole-brain AD',xlab='Corrected age (days)',sim.ci=TRUE,main='C')
plot_smooth(model.sex.ad,view="Corr_age",cond=list(Sex='F'),col='red',print.summary=FALSE,add=TRUE,sim.ci=TRUE)

plot_smooth(model.sex.rd, view="Corr_age", cond=list(Sex='M'), col='blue',print.summary=FALSE,hide.label=TRUE,ylab='Whole-brain RD',xlab='Corrected age (days)',sim.ci=TRUE,main='D')
plot_smooth(model.sex.rd,view="Corr_age",cond=list(Sex='F'),col='red',print.summary=FALSE,add=TRUE,sim.ci=TRUE)
dev.off()
```

## Growth rates
```{r}
model.fa.simple = gam(WB_fa.combat~s(Corr_age)+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
model.md.simple = gam(WB_md.combat~s(Corr_age)+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
model.ad.simple = gam(WB_ad.combat~s(Corr_age)+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
model.rd.simple = gam(WB_rd.combat~s(Corr_age)+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)

tiff('../Figures/GrowthCurveCorrectedAge.tiff',width=1080,height=320,pointsize=18)
par(mfrow=c(1,4))
plot_smooth(model.fa.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain FA",xlab="Corrected age (days)",rug=FALSE,main='A',col='slateblue',ylim=c(0.125,0.21))

# add subject-level data:
subjectList = unique(dat_harmonized$Unique_ID)
for (i in subjectList) {
  temp = dat_harmonized[dat_harmonized$Unique_ID==i,]
  lines(temp$WB_fa.combat~temp$Corr_age,col='gray',type='b',pch=20)
}
plot_smooth(model.fa.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain FA",xlab="Corrected age (days)",rug=FALSE,main='A',col='slateblue',ylim=c(0.125,0.21),add=TRUE)
  

plot_smooth(model.md.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain MD",xlab="Corrected age (days)",rug=FALSE,main='B',col='slateblue',ylim=c(0.875,1.35))
for (i in subjectList) {
  temp = dat_harmonized[dat_harmonized$Unique_ID==i,]
  lines(temp$WB_md.combat~temp$Corr_age,col='gray',type='b',pch=20)
}
plot_smooth(model.md.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain MD",xlab="Corrected age (days)",rug=FALSE,main='B',col='slateblue',add=TRUE)



plot_smooth(model.ad.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain AD",xlab="Corrected age (days)",rug=FALSE,main='C',col='slateblue',ylim=c(1.1,1.6))
for (i in subjectList) {
  temp = dat_harmonized[dat_harmonized$Unique_ID==i,]
  lines(temp$WB_ad.combat~temp$Corr_age,col='gray',type='b',pch=20)
}
plot_smooth(model.ad.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain AD",xlab="Corrected age (days)",rug=FALSE,main='C',col='slateblue',add=TRUE)


plot_smooth(model.rd.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain RD",xlab="Corrected age (days)",rug=FALSE,main='D',col='slateblue',ylim=c(0.8,1.3))
for (i in subjectList) {
  temp = dat_harmonized[dat_harmonized$Unique_ID==i,]
  lines(temp$WB_rd.combat~temp$Corr_age,col='gray',type='b',pch=20)
}

plot_smooth(model.rd.simple,view="Corr_age",print.summary=FALSE,hide.label=TRUE,sim.ci=TRUE,ylab="Whole-brain RD",xlab="Corrected  age (days)",rug=FALSE,main='D',col='slateblue',add=TRUE)

dev.off()



library(gratia)
library(ggplot2)


#dat_harmonized_GA40 = dat_harmonized
#dat_harmonized_GA40$GA = 40
#der.fa = derivatives(model0.corr.fa,newdata=dat_harmonized_GA40,term='ti(Corr_age)',type='central',interval='simultaneous',main='E',xlab='Corrected age (days)')
#draw(der.fa)

der.fa = derivatives(model.fa.simple,term='s(Corr_age)',type='central',interval='simultaneous',xlab='Corrected age (days)')

der.md = derivatives(model.md.simple,term='s(Corr_age)',type='central',interval='simultaneous',xlab='Corrected age (days)')

der.ad = derivatives(model.ad.simple,term='s(Corr_age)',type='central',interval='simultaneous',xlab='Corrected age (days)')

der.rd = derivatives(model.rd.simple,term='s(Corr_age)',type='central',interval='simultaneous',xlab='Corrected age (days)')

# p1=draw(der.fa,col='slateblue')+labs(title='E',x='Corrected age (days)',y='Derivative for FA')+theme_classic()
# p2=draw(der.md)+labs(title='F',x='Corrected age (days)',y='Derivative for MD')+theme_classic()
# p3=draw(der.ad)+labs(title='G',x='Corrected age (days)',y='Derivative for AD')+theme_classic()
# p4=draw(der.rd)+labs(title='H',x='Corrected age (days)',y='Derivative for RD')+theme_classic()

library(dplyr)
# make ggplot2 have centered and bold titles:
theme_update(plot.title = element_text(hjust = 0.5,face='bold')) # this didn't work for me

p1 =  der.fa %>%
   ggplot(aes(y = derivative, x = data)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.2, fill = "slateblue") +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "Rate of change FA",x='Corrected age (days)',title = "E") +
     theme_classic()+geom_hline(yintercept=0)+theme(plot.title = element_text(hjust = 0.5,face='bold'))

p2 =  der.md %>%
   ggplot(aes(y = derivative, x = data)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.2, fill = "slateblue") +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "Rate of change MD",x='Corrected age (days)',title = "F") +
     theme_classic()+geom_hline(yintercept=0)+theme(plot.title = element_text(hjust = 0.5,face='bold'))

p3 =  der.ad %>%
   ggplot(aes(y = derivative, x = data)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.2, fill = "slateblue") +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "Rate of change AD",x='Corrected age (days)',title = "G") +
     theme_classic()+geom_hline(yintercept=0)+theme(plot.title = element_text(hjust = 0.5,face='bold'))

p4 =  der.rd %>%
   ggplot(aes(y = derivative, x = data)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.2, fill = "slateblue") +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "Rate of change RD",x='Corrected age (days)',title = "H") +
     theme_classic()+geom_hline(yintercept=0)+theme(plot.title = element_text(hjust = 0.5,face='bold'))

library(gridExtra)
tiff('../Figures/GrowthRateCorrectedAge.tiff',width=1080,height=320,pointsize=18)
grid.arrange(p1,p2,p3,p4,nrow=1)
dev.off()
```

### Growth rates for text

```{r}
dat_harmonized_temp = dat_harmonized[rep(1,nrow(der.fa)),]
dat_harmonized_temp$Corr_age=der.fa$data

der.fa$predict = predict(model.fa.simple,newdata = dat_harmonized_temp)

# initial growth rate:
der.fa[1,]
  
  npredict = nrow(der.fa)
  # check incremented by 1:
  sum(abs(der.fa$data[2:npredict]-der.fa$data[1:(npredict-1)]-1)) # should be 0
    # approximately zero, looks good
  
  der.fa$first_order_approx=NA
  der.fa$first_order_approx[2:npredict] = der.fa$predict[2:npredict]-der.fa$predict[1:(npredict-1)]

  # eyeball if similar
  plot(der.fa$derivative~der.fa$first_order_approx)
  # looks good


  der.fa$fprimef = der.fa$derivative/der.fa$predict
  
# Calculate the percent change 
der.fa$perc_change = NA
der.fa$perc_change[1:(npredict-1)] = 100*der.fa$first_order_approx[2:npredict]/der.fa$predict[1:(npredict-1)]
  
der.md$predict = predict(model.md.simple,newdata = dat_harmonized_temp)
der.md$first_order_approx=NA
der.md$first_order_approx[2:npredict] = der.md$predict[2:npredict]-der.md$predict[1:(npredict-1)]  
der.md$perc_change = NA
der.md$perc_change[1:(npredict-1)] = 100*der.md$first_order_approx[2:npredict]/der.md$predict[1:(npredict-1)]

der.ad$predict = predict(model.ad.simple,newdata = dat_harmonized_temp)
der.ad$first_order_approx=NA
der.ad$first_order_approx[2:npredict] = der.ad$predict[2:npredict]-der.ad$predict[1:(npredict-1)]  
der.ad$perc_change = NA
der.ad$perc_change[1:(npredict-1)] = 100*der.ad$first_order_approx[2:npredict]/der.ad$predict[1:(npredict-1)]

der.rd$predict = predict(model.rd.simple,newdata = dat_harmonized_temp)
der.rd$first_order_approx=NA
der.rd$first_order_approx[2:npredict] = der.rd$predict[2:npredict]-der.rd$predict[1:(npredict-1)]  
der.rd$perc_change = NA
der.rd$perc_change[1:(npredict-1)] = 100*der.rd$first_order_approx[2:npredict]/der.rd$predict[1:(npredict-1)]

# initial growth rate percentages:
der.fa$perc_change[1]
der.md$perc_change[1]
der.ad$perc_change[1]
der.rd$perc_change[1]

# fa, growth rate at 100 days where starts to flatten
der.fa[99,c('data','derivative','perc_change')]
der.fa$data[149]
der.fa$data[199]
mean(der.fa$perc_change[149:199])

der.md$perc_change[1]
der.ad$perc_change[1]
der.rd$perc_change[1]

der.md[69,c('data','perc_change')]
der.ad[69,c('data','perc_change')]
der.rd[69,c('data','perc_change')]

# average growth rate 160 to 200
der.md[159,c('data','perc_change')]
mean(der.md$perc_change[159:199])
mean(der.ad$perc_change[159:199])
mean(der.rd$perc_change[159:199])


```

### Fold analysis:
```{r}
der.fa.6mos = der.fa$derivative[round(der.fa$data)==180]
der.fa$fold = der.fa$derivative/der.fa.6mos

der.md.6mos = der.md$derivative[round(der.md$data)==180]
der.md$fold = der.md$derivative/der.md.6mos

der.ad.6mos = der.ad$derivative[round(der.ad$data)==180]
der.ad$fold = der.ad$derivative/der.ad.6mos

der.rd.6mos = der.rd$derivative[round(der.rd$data)==180]
der.rd$fold = der.rd$derivative/der.rd.6mos

der.fa$fold[1]
der.ad$fold[1]
der.rd$fold[1]
der.md$fold[1]

p1 =  der.fa %>%
   ggplot(aes(y = fold, x = data)) +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "(FA Rate of Change)/(FA Rate of Change at 6 mos)",x='Corrected age (days)',title = "A") +
     theme_classic()+geom_vline(xintercept=180)+theme(plot.title = element_text(hjust = 0.5,face='bold'))+ylim(0.5,7.25)

p2 =  der.md %>%
   ggplot(aes(y = fold, x = data)) +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "(MD Rate of Change)/(MD Rate of Change at 6 mos)",x='Corrected age (days)',title = "B") +
     theme_classic()+geom_vline(xintercept=180)+theme(plot.title = element_text(hjust = 0.5,face='bold'))


p3 =  der.ad %>%
   ggplot(aes(y = fold, x = data)) +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "(AD Rate of Change)/(AD Rate of Change at 6 mos)",x='Corrected age (days)',title = "C") +
     theme_classic()+geom_vline(xintercept=180)+theme(plot.title = element_text(hjust = 0.5,face='bold'))


p4 =  der.rd %>%
   ggplot(aes(y = fold, x = data)) +
     geom_line(colour = "slateblue", size = 1.5) +
     labs(y = "(RD Rate of Change)/(RD Rate of Change at 6 mos)",x='Corrected age (days)',title = "D") +
     theme_classic()+geom_vline(xintercept=180)+theme(plot.title = element_text(hjust = 0.5,face='bold'))

tiff('../Figures/FoldAnalysisWB.tiff',width=1080,height=320,pointsize=18)
grid.arrange(p1,p2,p3,p4,nrow=1)
dev.off()

```
