---
title: 'White Matter Development: Analysis'
author: "Benjamin Risk"
date: "4/27/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mgcv)
library(itsadug)
dat_harmonized = read.csv('dat_harmonized.csv',stringsAsFactors = TRUE)
```


Fx -- Fornix (limbic, output of hippocampus)

IFOF -- inferior-fronto-occipital-fasciculus (associative, connects occipital cortex to frontal lobe)

CCb -- body of corpus callosum (commissural, connects hemispheres)

PT -- pyramidal tract includes pyramidal neurons connecting motor neurons to brain stem or spinal cord (corticospinal and corticobulbar tracts)


### FA

This function extracts the largest age at which there is a significant difference:
```{r}
# note: full term is 37
maxChronAge = function(GAlow,GAhigh=40,model,ylab,less=TRUE,print.plot=FALSE) {
  a=plot_diff(model,view='chor_age',comp=list(GA=c(GAlow,GAhigh)),sim.ci = TRUE, hide.label=TRUE,xlab='Chronological age (days)',ylab=ylab,plot=print.plot)
  if(less) {
  max(a$chor_age[a$est+a$sim.CI<0])
  } else {
    max(a$chor_age[a$est-a$sim.CI>0])
  }
}

# Fx and Pt can have opposite trend, become more dissimilar with age
minChronAge = function(GAlow,GAhigh=40,model,ylab,less=TRUE,print.plot=FALSE) {
  a=plot_diff(model,view='chor_age',comp=list(GA=c(GAlow,GAhigh)),sim.ci = TRUE, hide.label=TRUE,xlab='Chronological age (days)',ylab=ylab,plot=print.plot)
  if(less) {
  min(a$chor_age[a$est+a$sim.CI<0])
  } else {
    min(a$chor_age[a$est-a$sim.CI>0])
  }
}

```

The syntax used here is described as the functional anova in the supporting R documentation for ?ti(). 
```{r,error=TRUE}

# Create plots at different gestational ages:
my.function.ga <- function(dti.tract, my.ylim = c(0.1,0.3), my.ylab,  my.main="", estMaxChronAge=TRUE,estMinChronAge=FALSE,print.legend=FALSE,less=TRUE) {
    model0 = gam(dti.tract~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
  # step into function to examine model diagnostics:
 # gam.check(model0)
  print(summary(model0))

  # vis.gam(model0,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='FA')
  set.seed(123)

  #tiff('../Figures/DiffCurves_FA.tiff',width=720,height=480,pointsize=20)
  if (estMaxChronAge) {
      # par(mfrow=c(2,3))
      (ga.35=maxChronAge(GAlow=35,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.36=maxChronAge(GAlow=36,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.37=maxChronAge(GAlow=37,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.38=maxChronAge(GAlow=38,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.39=maxChronAge(GAlow=39,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.41=maxChronAge(GAlow=40,GAhigh=41,model=model0,ylab='Est. difference',less=less))
      #dev.off()
  }
  
  if (estMinChronAge) {
      # par(mfrow=c(2,3))
      (ga.35=minChronAge(GAlow=35,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.36=minChronAge(GAlow=36,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.37=minChronAge(GAlow=37,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.38=minChronAge(GAlow=38,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.39=minChronAge(GAlow=39,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.41=minChronAge(GAlow=40,GAhigh=41,model=model0,ylab='Est. difference',less=less))
      #dev.off()
  }
  
  cColors = viridis::plasma(15)
  
    model0.plot = plot_smooth(model0, view="chor_age",cond=list(GA=35),col=cColors[1],rug=FALSE,ylim=my.ylim,lwd=2,se=0,ylab=my.ylab,xlab='Chronological age (days)',hide.label=TRUE,main=my.main,xlim=c(10,202),print.summary=FALSE)
  plot_smooth(model0, view="chor_age",cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2,se=0,print.summary=FALSE)
  plot_smooth(model0, view="chor_age",  cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2,se=0,print.summary=FALSE)
  plot_smooth(model0, view="chor_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE)
  plot_smooth(model0, view="chor_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE)
  plot_smooth(model0, view="chor_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE)
  plot_smooth(model0, view="chor_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE)
  
  if(print.legend==TRUE) {
  legend(x='bottomright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1)
  }
  if (estMaxChronAge | estMinChronAge) {
    abline(v=ga.41,col=cColors[13],lwd=2,lty=2)
    abline(v=ga.39,col=cColors[9],lwd=2,lty=2)
    abline(v=ga.38,col=cColors[7],lwd=2,lty=2)
    abline(v=ga.37,col=cColors[5],lwd=2,lty=2)
    abline(v=ga.36,col=cColors[3],lwd=2,lty=2)
    abline(v=ga.35,col=cColors[1],lwd=2,lty=2)
  }
  my.plot = recordPlot()
  my.plot
}
```


### FA

```{r}
tiff('../Figures/TractSpecific_GestandCorrectedAgeFA.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

# Note on interaction effects:
# Fx p=0.69, CCs p=0.10; ILF p==0.41, AF p=0.13
fx.fa = my.function.ga(dti.tract = dat_harmonized$Fx_fa.combat, my.ylab = "FX FA")

ci.fa = my.function.ga(dti.tract = dat_harmonized$Ci_fa.combat, my.ylab = "Ci FA")

ccg.fa = my.function.ga(dti.tract = dat_harmonized$CCg_fa.combat, my.ylab = "CCg FA")

ccb.fa = my.function.ga(dti.tract = dat_harmonized$CCb_fa.combat, my.ylab="CCb FA")

ccs.fa = my.function.ga(dti.tract = dat_harmonized$CCs_fa.combat, my.ylab="CCs FA")

ilf.fa = my.function.ga(dti.tract = dat_harmonized$ILF_fa.combat, my.ylab="ILF FA")

ifof.fa = my.function.ga(dti.tract = dat_harmonized$IFOF_fa.combat, my.ylab="IFOF FA")

af.fa = my.function.ga(dti.tract = dat_harmonized$AF_fa.combat, my.ylab = "AF FA")

uf.fa = my.function.ga(dti.tract = dat_harmonized$UF_fa.combat, my.ylab="UF FA")

atr.fa = my.function.ga(dti.tract = dat_harmonized$ATR_fa.combat, my.ylab="ATR FA")

pt.fa = my.function.ga(dti.tract = dat_harmonized$PT_fa.combat, my.ylab="PT FA",print.legend = FALSE)

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=0.5)

dev.off()

```

### MD

```{r}
tiff('../Figures/TractSpecific_GestandCorrectedAgeMD.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

fx = my.function.ga(dti.tract = dat_harmonized$Fx_md.combat, my.ylab = "FX MD",less=FALSE,my.ylim = c(0.9,1.5),estMinChronAge = FALSE,estMaxChronAge=FALSE)

ci = my.function.ga(dti.tract = dat_harmonized$Ci_md.combat, my.ylab = "Ci MD",my.ylim = c(0.9,1.5),less=FALSE)

ccg = my.function.ga(dti.tract = dat_harmonized$CCg_md.combat, my.ylab = "CCg MD",my.ylim = c(0.9,1.5),less=FALSE)

ccb = my.function.ga(dti.tract = dat_harmonized$CCb_md.combat, my.ylab="CCb MD",my.ylim = c(0.9,1.5),less=FALSE)

ccs = my.function.ga(dti.tract = dat_harmonized$CCs_md.combat, my.ylab="CCs MD",less=FALSE,my.ylim = c(0.9,1.5))

ilf = my.function.ga(dti.tract = dat_harmonized$ILF_md.combat, my.ylab="ILF MD",less=FALSE,my.ylim = c(0.9,1.5))

ifof = my.function.ga(dti.tract = dat_harmonized$IFOF_md.combat, my.ylab="IFOF MD",less=FALSE,my.ylim=c(0.9,1.5))

af = my.function.ga(dti.tract = dat_harmonized$AF_md.combat, my.ylab = "AF MD",my.ylim=c(0.9,1.5),less=FALSE,estMaxChronAge=FALSE)

uf = my.function.ga(dti.tract = dat_harmonized$UF_md.combat, my.ylab="UF MD",less=FALSE,my.ylim=c(0.9,1.5))

atr = my.function.ga(dti.tract = dat_harmonized$ATR_md.combat, my.ylab="ATR MD",less=FALSE,my.ylim=c(0.9,1.5))

pt = my.function.ga(dti.tract = dat_harmonized$PT_md.combat, my.ylab="PT MD",print.legend = FALSE,less=FALSE,estMinChronAge=FALSE,estMaxChronAge=FALSE,my.ylim=c(0.9,1.5))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```
For FX and PT, there is trend for the differences to increase over time, and chronological age less than the dashed lines do not significantly differ between the denoted GA and GA=40, while chronological age greater than the dashed lines significantly differ.


### AD

```{r}
tiff('../Figures/TractSpecific_GestandCorrectedAgeAD.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

fx = my.function.ga(dti.tract = dat_harmonized$Fx_ad.combat, my.ylab = "FX AD",less=FALSE,my.ylim = c(1.1,1.7),estMinChronAge = FALSE,estMaxChronAge=FALSE)

ci = my.function.ga(dti.tract = dat_harmonized$Ci_ad.combat, my.ylab = "Ci AD",my.ylim = c(1.1,1.7),less=FALSE)

ccg = my.function.ga(dti.tract = dat_harmonized$CCg_ad.combat, my.ylab = "CCg AD",my.ylim = c(1.1,1.7),less=FALSE)

ccb = my.function.ga(dti.tract = dat_harmonized$CCb_ad.combat, my.ylab="CCb AD",my.ylim = c(1.1,1.7),less=FALSE)

ccs = my.function.ga(dti.tract = dat_harmonized$CCs_ad.combat, my.ylab="CCs AD",less=FALSE,my.ylim = c(1.1,1.7))

ilf = my.function.ga(dti.tract = dat_harmonized$ILF_ad.combat, my.ylab="ILF AD",less=FALSE,my.ylim = c(1.1,1.7))

ifof = my.function.ga(dti.tract = dat_harmonized$IFOF_ad.combat, my.ylab="IFOF AD",less=FALSE,my.ylim = c(1.1,1.7))

af = my.function.ga(dti.tract = dat_harmonized$AF_ad.combat, my.ylab = "AF AD",my.ylim = c(1.1,1.7),less=FALSE,estMaxChronAge = FALSE)

uf = my.function.ga(dti.tract = dat_harmonized$UF_ad.combat, my.ylab="UF AD",less=FALSE,my.ylim = c(1.1,1.7))

atr = my.function.ga(dti.tract = dat_harmonized$ATR_ad.combat, my.ylab="ATR AD",less=FALSE,my.ylim = c(1.1,1.7))

pt = my.function.ga(dti.tract = dat_harmonized$PT_ad.combat, my.ylab="PT AD",print.legend = FALSE,less=FALSE,estMinChronAge=FALSE,estMaxChronAge=FALSE,my.ylim = c(1.1,1.7))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```
AF is strange: small region over which there are significant differences, approximately 50-70 days. For clarify, don't plus the significance lines. 


### RD

```{r}
tiff('../Figures/TractSpecific_GestandCorrectedAgeRD.tiff',width=1080,height=1280,pointsize=18)
par(mfrow=c(4,3))

fx = my.function.ga(dti.tract = dat_harmonized$Fx_rd.combat, my.ylab = "FX RD",less=FALSE,my.ylim = c(0.8,1.4),estMinChronAge = FALSE,estMaxChronAge=FALSE)

ci = my.function.ga(dti.tract = dat_harmonized$Ci_rd.combat, my.ylab = "Ci RD",my.ylim = c(0.8,1.4),less=FALSE)

ccg = my.function.ga(dti.tract = dat_harmonized$CCg_rd.combat, my.ylab = "CCg RD",my.ylim = c(0.8,1.4),less=FALSE)

ccb = my.function.ga(dti.tract = dat_harmonized$CCb_rd.combat, my.ylab="CCb RD",my.ylim = c(0.8,1.4),less=FALSE)

ccs = my.function.ga(dti.tract = dat_harmonized$CCs_rd.combat, my.ylab="CCs RD",less=FALSE,my.ylim = c(0.8,1.4))

ilf = my.function.ga(dti.tract = dat_harmonized$ILF_rd.combat, my.ylab="ILF RD",less=FALSE,my.ylim = c(0.8,1.4))

ifof = my.function.ga(dti.tract = dat_harmonized$IFOF_rd.combat, my.ylab="IFOF RD",less=FALSE,my.ylim = c(0.8,1.4))

af = my.function.ga(dti.tract = dat_harmonized$AF_rd.combat, my.ylab = "AF RD",my.ylim = c(0.8,1.4),less=FALSE)

uf = my.function.ga(dti.tract = dat_harmonized$UF_rd.combat, my.ylab="UF RD",less=FALSE,my.ylim = c(0.8,1.4))

atr = my.function.ga(dti.tract = dat_harmonized$ATR_rd.combat, my.ylab="ATR RD",less=FALSE,my.ylim = c(0.8,1.4))

pt = my.function.ga(dti.tract = dat_harmonized$PT_rd.combat, my.ylab="PT RD",print.legend = FALSE,less=FALSE,estMinChronAge=FALSE,estMaxChronAge=FALSE,my.ylim = c(0.8,1.4))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```






```{r}

tiff('../Figures/TractSpecific_GestandCorrectedAgeRD.tiff',width=1080,height=1280,pointsize=18)
par(mfrow=c(4,3))

#pdf(file='../Figures/TractSpecific_GestandCorrectedAgeRD.pdf')

fx = my.function.ga(dti.tract = dat_harmonized$Fx_rd.combat, my.ylab = "FX RD",less=FALSE,my.ylim = c(0.8,1.4),estMinChronAge = FALSE,estMaxChronAge=FALSE)

ci = my.function.ga(dti.tract = dat_harmonized$Ci_rd.combat, my.ylab = "Ci RD",my.ylim = c(0.8,1.4),less=FALSE)

ccg = my.function.ga(dti.tract = dat_harmonized$CCg_rd.combat, my.ylab = "CCg RD",my.ylim = c(0.8,1.4),less=FALSE)

ccb = my.function.ga(dti.tract = dat_harmonized$CCb_rd.combat, my.ylab="CCb RD",my.ylim = c(0.8,1.4),less=FALSE)

ccs = my.function.ga(dti.tract = dat_harmonized$CCs_rd.combat, my.ylab="CCs RD",less=FALSE,my.ylim = c(0.8,1.4))

ilf = my.function.ga(dti.tract = dat_harmonized$ILF_rd.combat, my.ylab="ILF RD",less=FALSE,my.ylim = c(0.8,1.4))

ifof = my.function.ga(dti.tract = dat_harmonized$IFOF_rd.combat, my.ylab="IFOF RD",less=FALSE,my.ylim = c(0.8,1.4))

af = my.function.ga(dti.tract = dat_harmonized$AF_rd.combat, my.ylab = "AF RD",my.ylim = c(0.8,1.4),less=FALSE)

uf = my.function.ga(dti.tract = dat_harmonized$UF_rd.combat, my.ylab="UF RD",less=FALSE,my.ylim = c(0.8,1.4))

atr = my.function.ga(dti.tract = dat_harmonized$ATR_rd.combat, my.ylab="ATR RD",less=FALSE,my.ylim = c(0.8,1.4))

pt = my.function.ga(dti.tract = dat_harmonized$PT_rd.combat, my.ylab="PT RD",print.legend = FALSE,less=FALSE,estMinChronAge=FALSE,estMaxChronAge=FALSE,my.ylim = c(0.8,1.4))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()

```