---
title: 'White Matter Develoment: Analysis'
author: "Benjamin Risk"
date: "4/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Examine scanner effects
We examine whether combat reduced batch effects. This code also allows a comparison of the curves in the trio and prisma scanners.

```{r}
library(mgcv)
library(itsadug)

examine_combat_corrage = function(feature,feature.combat,dataset,legendlocation='bottomright',ylab,xlab) {
 
#  model0 = gam(feature~Scanner+s(Corr_age,by=Scanner,bs='bs')+Sex+s(Unique_ID,bs='re'),method='REML',data=dataset)

#  model0.harm = gam(feature.combat~Scanner+s(Corr_age,by=Scanner,bs='bs')+Sex+s(Unique_ID,bs='re')+Scanner,method='REML',data=dataset)
# Edited 12 June 2022
    model0 = gam(feature~Scanner+s(Corr_age,by=Scanner)+Sex+s(Unique_ID,bs='re'),method='REML',data=dataset)

  model0.harm = gam(feature.combat~Scanner+s(Corr_age,by=Scanner)+Sex+s(Unique_ID,bs='re')+Scanner,method='REML',data=dataset)

   
   par(mfrow=c(2,3))
  b=plot_smooth(model0,view="Corr_age",cond=list(Scanner='trio'),col='red',print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', no harmonize'),xlab=xlab)
  
  plot_smooth(model0, view="Corr_age", cond=list(Scanner='prisma'), add=TRUE, col='blue',print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', no harmonize'),xlab=xlab)
    legend(x=legendlocation,legend=c('trio','prisma'),lty=1,lwd=2,col=c('red','blue'),bty='n',cex=1)
  plot_diff(model0,view='Corr_age',comp=list(Scanner=c('trio','prisma')),print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', no harmonize'),xlab=xlab)
  boxplot(resid(model0)~dataset$Scanner,ylab='residual, no harmonize',xlab='Scanner')
  

  plot_smooth(model0.harm,view="Corr_age",cond=list(Scanner='trio'),col='red',print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', harmonized'),xlab=xlab)
  plot_smooth(model0.harm, view="Corr_age", cond=list(Scanner='prisma'), add=TRUE, col='blue',print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', harmonized'),xlab=xlab)
    legend(x=legendlocation,legend=c('trio','prisma'),lty=1,lwd=2,col=c('red','blue'),bty='n',cex=1)
  plot_diff(model0.harm,view='Corr_age',comp=list(Scanner=c('trio','prisma')),print.summary=FALSE,sim.ci=TRUE,hide.label=TRUE,ylab=paste0(ylab,', harmonized'),xlab=xlab)
  boxplot(resid(model0.harm)~dataset$Scanner,ylab='residual, harmonized',xlab='Scanner')
}
```

Examine the whole brain curves:
```{r}
dat_harmonized = read.csv('dat_harmonized.csv',stringsAsFactors = TRUE)

tiff(file='../Figures/WholeBrainFA_Combat.tiff',width=720,height=480,pointsize=20)
examine_combat_corrage(dat_harmonized$WB_fa,dat_harmonized$WB_fa.combat,dat_harmonized,xlab='Corrected age (days)',ylab='fa')
dev.off()

tiff(file='../Figures/WholeBrainMD_Combat.tiff',width=720,height=480,pointsize=20)
examine_combat_corrage(dat_harmonized$WB_md,dat_harmonized$WB_md.combat,dat_harmonized,xlab='Corrected age (days)',ylab='md',legendlocation='topright')
dev.off()

tiff(file='../Figures/WholeBrainAD_Combat.tiff',width=720,height=480,pointsize=20)
examine_combat_corrage(dat_harmonized$WB_ad,dat_harmonized$WB_ad.combat,dat_harmonized,xlab='Corrected age (days)',ylab='ad',legendlocation='topright')
dev.off()

tiff(file='../Figures/WholeBrainRD_Combat.tiff',width=720,height=480,pointsize=20)
examine_combat_corrage(dat_harmonized$WB_rd,dat_harmonized$WB_rd.combat,dat_harmonized,xlab='Corrected age (days)',ylab='rd',legendlocation='topright')
dev.off()

```
Combat is helping to remove the shift in the mean. I don't see a large change in the variances, though, as indicated by the IQR, but looks okay for further analysis.


Examine a few of the trajectories in different ROIs. These results are not in the current version of the manuscript:
```{r,eval=FALSE}
examine_combat(dat_harmonized$AF_fa,dat_harmonized$AF_fa.combat,dat_harmonized,xlab='Corrected Age (days)',ylab='fa')

examine_combat(dat_harmonized$CCg_fa,dat_harmonized$CCg_fa.combat,dat_harmonized,xlab='Corrected Age (days)',ylab='fa')

examine_combat(dat_harmonized$CCg_md,dat_harmonized$CCg_md.combat,dat_harmonized,xlab='Corrected Age (days)',ylab='fa')

examine_combat(dat_harmonized$PT_rd,dat_harmonized$PT_rd.combat,dat_harmonized,xlab='Corrected Age (days)',ylab='fa')

```
