---
title: 'White Matter Development: Analysis'
author: "Benjamin Risk"
date: "5/27/2024"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mgcv)
library(itsadug)
dat_harmonized = read.csv('dat_harmonized.csv',stringsAsFactors = TRUE)
```


Fx -- Fornix (limbic, output of hippocampus)

IFOF -- inferior-fronto-occipital-fasciculus (associative, connects occipital cortex to frontal lobe)

CCb -- body of corpus callosum (commissural, connects hemispheres)

PT -- pyramidal tract includes pyramidal neurons connecting motor neurons to brain stem or spinal cord (corticospinal and corticobulbar tracts)

```{r}
#Function for examining whether there is an interaction between GA and corrected age
### FA
maxCorrAge = function(GAlow,GAhigh=40,model,ylab,less=TRUE,print.plot=FALSE) {
  a=plot_diff(model,view='Corr_age',comp=list(GA=c(GAlow,GAhigh)),sim.ci = TRUE, hide.label=TRUE,xlab='Corrected age (days)',ylab=ylab,plot=print.plot)
  if(less) {
  max(a$Corr_age[a$est+a$sim.CI<0])
  } else {
    max(a$Corr_age[a$est-a$sim.CI>0])
  }
}
minCorrAge = function(GAlow,GAhigh=40,model,ylab,less=TRUE,print.plot=FALSE) {
  a=plot_diff(model,view='Corr_age',comp=list(GA=c(GAlow,GAhigh)),sim.ci = TRUE, hide.label=TRUE,xlab='Corrected age (days)',ylab=ylab,plot=print.plot)
  if(less) {
  min(a$Corr_age[a$est+a$sim.CI<0])
  } else {
    min(a$Corr_age[a$est-a$sim.CI>0])
  }
}

# Create plots at different gestational ages:
my.function.ga.corr <- function(dti.tract, my.ylim = c(0.1,0.3), my.ylab,  my.main="", estMaxCorrAge=TRUE,estMinCorrAge=FALSE,print.legend=FALSE,less=TRUE) {
    model0 = gam(dti.tract~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
  # step into function to examine model diagnostics:
 # gam.check(model0)
  print(summary(model0))

  # vis.gam(model0,view=c('chor_age','GA'),theta=-45,ticktype='detailed',xlab='Chron Age',ylab='Gest Age',zlab='FA')
  set.seed(123)

  #tiff('../Figures/DiffCurves_FA.tiff',width=720,height=480,pointsize=20)
  if (estMaxCorrAge) {
      # par(mfrow=c(2,3))
      (ga.35=maxCorrAge(GAlow=35,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.36=maxCorrAge(GAlow=36,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.37=maxCorrAge(GAlow=37,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.38=maxCorrAge(GAlow=38,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.39=maxCorrAge(GAlow=39,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.41=maxCorrAge(GAlow=40,GAhigh=41,model=model0,ylab='Est. difference',less=less))
      #dev.off()
  }
  
  if (estMinCorrAge) {
      # par(mfrow=c(2,3))
      (ga.35=minCorrAge(GAlow=35,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.36=minCorrAge(GAlow=36,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.37=minCorrAge(GAlow=37,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.38=minCorrAge(GAlow=38,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.39=minCorrAge(GAlow=39,GAhigh=40,model=model0,ylab='Est. difference',less=less))
      (ga.41=minCorrAge(GAlow=40,GAhigh=41,model=model0,ylab='Est. difference',less=less))
      #dev.off()
  }
  
  cColors = viridis::plasma(15)
  
    model0.plot = plot_smooth(model0, view="Corr_age",cond=list(GA=35),col=cColors[1],rug=FALSE,ylim=my.ylim,lwd=2,se=0,ylab=my.ylab,xlab='Corrected age (days)',hide.label=TRUE,main=my.main,xlim=c(2,201),print.summary=FALSE)
  plot_smooth(model0, view="Corr_age",cond=list(GA=36),col=cColors[3],add=TRUE,rug=FALSE,lwd=2,se=0,print.summary=FALSE)
  plot_smooth(model0, view="Corr_age",  cond=list(GA=37),col=cColors[5],add=TRUE,rug=FALSE,lwd=2,se=0,print.summary=FALSE,xlim=c(2,201))
  plot_smooth(model0, view="Corr_age", se=0, cond=list(GA=38),col=cColors[7],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE,xlim=c(2,201))
  plot_smooth(model0, view="Corr_age", se=0, cond=list(GA=39),col=cColors[9],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE,xlim=c(2,201))
  plot_smooth(model0, view="Corr_age", se=0, cond=list(GA=40),col=cColors[11],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE,xlim=c(2,201))
  plot_smooth(model0, view="Corr_age", se=0, cond=list(GA=41),col=cColors[13],add=TRUE,rug=FALSE,lwd=2,print.summary=FALSE,xlim=c(2,201))
  
  if(print.legend==TRUE) {
  legend(x='bottomright',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1)
  }
  if (estMaxCorrAge | estMinCorrAge) {
    abline(v=ga.41,col=cColors[13],lwd=2,lty=2)
    abline(v=ga.39,col=cColors[9],lwd=2,lty=2)
    abline(v=ga.38,col=cColors[7],lwd=2,lty=2)
    abline(v=ga.37,col=cColors[5],lwd=2,lty=2)
    abline(v=ga.36,col=cColors[3],lwd=2,lty=2)
    abline(v=ga.35,col=cColors[1],lwd=2,lty=2)
  }
  my.plot = recordPlot()
  my.plot
}


```


```{r}
tiff('../Figures/TractSpecific_GestCorrectedAgeFA.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

fx.fa = my.function.ga.corr(dti.tract = dat_harmonized$Fx_fa.combat, my.ylab = "FX FA")

# check if there is a difference for entire time range or no difference anywhere:
#model0 = gam(Fx_fa.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
#plot_diff(model0,view='chor_age',comp=list(GA=c(35,40)),sim.ci = TRUE, hide.label=TRUE,xlab='Chronological age (days)')


ci.fa = my.function.ga.corr(dti.tract = dat_harmonized$Ci_fa.combat, my.ylab = "Ci FA")

ccg.fa = my.function.ga.corr(dti.tract = dat_harmonized$CCg_fa.combat, my.ylab = "CCg FA")

ccb.fa = my.function.ga.corr(dti.tract = dat_harmonized$CCb_fa.combat, my.ylab="CCb FA")

ccs.fa = my.function.ga.corr(dti.tract = dat_harmonized$CCs_fa.combat, my.ylab="CCs FA")

ilf.fa = my.function.ga.corr(dti.tract = dat_harmonized$ILF_fa.combat, my.ylab="ILF FA")

#model0 = gam(ILF_fa.combat~ti(chor_age,k=9)+ti(GA,k=5)+ti(chor_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
#plot_diff(model0,view='chor_age',comp=list(GA=c(35,40)),sim.ci = TRUE, hide.label=TRUE,xlab='Chronological age (days)')



ifof.fa = my.function.ga.corr(dti.tract = dat_harmonized$IFOF_fa.combat, my.ylab="IFOF FA")

af.fa = my.function.ga.corr(dti.tract = dat_harmonized$AF_fa.combat, my.ylab = "AF FA")

uf.fa = my.function.ga.corr(dti.tract = dat_harmonized$UF_fa.combat, my.ylab="UF FA")

atr.fa = my.function.ga.corr(dti.tract = dat_harmonized$ATR_fa.combat, my.ylab="ATR FA")

pt.fa = my.function.ga.corr(dti.tract = dat_harmonized$PT_fa.combat, my.ylab="PT FA",print.legend = FALSE)

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()

```

### MD

```{r}
tiff('../Figures/TractSpecific_GestCorrectedAgeMD.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

fx = my.function.ga.corr(dti.tract = dat_harmonized$Fx_md.combat, my.ylab = "FX MD",less=FALSE,my.ylim = c(0.9,1.5),estMinCorrAge = FALSE,estMaxCorrAge=FALSE)

ci = my.function.ga.corr(dti.tract = dat_harmonized$Ci_md.combat, my.ylab = "Ci MD",my.ylim = c(0.9,1.5),less=FALSE)

ccg = my.function.ga.corr(dti.tract = dat_harmonized$CCg_md.combat, my.ylab = "CCg MD",my.ylim = c(0.9,1.5),less=FALSE)

ccb = my.function.ga.corr(dti.tract = dat_harmonized$CCb_md.combat, my.ylab="CCb MD",my.ylim = c(0.9,1.5),less=FALSE)

ccs = my.function.ga.corr(dti.tract = dat_harmonized$CCs_md.combat, my.ylab="CCs MD",less=FALSE,my.ylim = c(0.9,1.5))

ilf = my.function.ga.corr(dti.tract = dat_harmonized$ILF_md.combat, my.ylab="ILF MD",less=FALSE,my.ylim = c(0.9,1.5))

ifof = my.function.ga.corr(dti.tract = dat_harmonized$IFOF_md.combat, my.ylab="IFOF MD",less=FALSE,my.ylim=c(0.9,1.5))

af = my.function.ga.corr(dti.tract = dat_harmonized$AF_md.combat, my.ylab = "AF MD",my.ylim=c(0.9,1.5),less=FALSE,estMaxCorrAge=FALSE)

uf = my.function.ga.corr(dti.tract = dat_harmonized$UF_md.combat, my.ylab="UF MD",less=FALSE,my.ylim=c(0.9,1.5))

atr = my.function.ga.corr(dti.tract = dat_harmonized$ATR_md.combat, my.ylab="ATR MD",less=FALSE,my.ylim=c(0.9,1.5))

pt = my.function.ga.corr(dti.tract = dat_harmonized$PT_md.combat, my.ylab="PT MD",print.legend = FALSE,less=FALSE,estMinCorrAge=FALSE,estMaxCorrAge=FALSE,my.ylim=c(0.9,1.5))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```


### AD

```{r}
tiff('../Figures/TractSpecific_GestCorrectedAgeAD.tiff',width=1080,height=1280,pointsize=18)

par(mfrow=c(4,3))

fx = my.function.ga.corr(dti.tract = dat_harmonized$Fx_ad.combat, my.ylab = "FX AD",less=FALSE,my.ylim = c(1.1,1.7),estMinCorrAge = FALSE,estMaxCorrAge=FALSE)

ci = my.function.ga.corr(dti.tract = dat_harmonized$Ci_ad.combat, my.ylab = "Ci AD",my.ylim = c(1.1,1.7),less=FALSE)

ccg = my.function.ga.corr(dti.tract = dat_harmonized$CCg_ad.combat, my.ylab = "CCg AD",my.ylim = c(1.1,1.7),less=FALSE)

ccb = my.function.ga.corr(dti.tract = dat_harmonized$CCb_ad.combat, my.ylab="CCb AD",my.ylim = c(1.1,1.7),less=FALSE)

ccs = my.function.ga.corr(dti.tract = dat_harmonized$CCs_ad.combat, my.ylab="CCs AD",less=FALSE,my.ylim = c(1.1,1.7))

ilf = my.function.ga.corr(dti.tract = dat_harmonized$ILF_ad.combat, my.ylab="ILF AD",less=FALSE,my.ylim = c(1.1,1.7))

ifof = my.function.ga.corr(dti.tract = dat_harmonized$IFOF_ad.combat, my.ylab="IFOF AD",less=FALSE,my.ylim = c(1.1,1.7))

af = my.function.ga.corr(dti.tract = dat_harmonized$AF_ad.combat, my.ylab = "AF AD",my.ylim = c(1.1,1.7),less=FALSE,estMaxCorrAge = FALSE)

uf = my.function.ga.corr(dti.tract = dat_harmonized$UF_ad.combat, my.ylab="UF AD",less=FALSE,my.ylim = c(1.1,1.7))

atr = my.function.ga.corr(dti.tract = dat_harmonized$ATR_ad.combat, my.ylab="ATR AD",less=FALSE,my.ylim = c(1.1,1.7))

pt = my.function.ga.corr(dti.tract = dat_harmonized$PT_ad.combat, my.ylab="PT AD",print.legend = FALSE,less=FALSE,estMinCorrAge=FALSE,estMaxCorrAge=FALSE,my.ylim = c(1.1,1.7))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```
AF is strange: small region over which there are significant differences, approximately 50-70 days. For clarify, don't plus the significance lines. 


### RD

```{r}
tiff('../Figures/TractSpecific_GestCorrectedAgeRD.tiff',width=1080,height=1280,pointsize=18)
par(mfrow=c(4,3))

fx = my.function.ga.corr(dti.tract = dat_harmonized$Fx_rd.combat, my.ylab = "FX RD",less=FALSE,my.ylim = c(0.8,1.4),estMinCorrAge = FALSE,estMaxCorrAge=FALSE)

ci = my.function.ga.corr(dti.tract = dat_harmonized$Ci_rd.combat, my.ylab = "Ci RD",my.ylim = c(0.8,1.4),less=FALSE)

ccg = my.function.ga.corr(dti.tract = dat_harmonized$CCg_rd.combat, my.ylab = "CCg RD",my.ylim = c(0.8,1.4),less=FALSE)

ccb = my.function.ga.corr(dti.tract = dat_harmonized$CCb_rd.combat, my.ylab="CCb RD",my.ylim = c(0.8,1.4),less=FALSE)

ccs = my.function.ga.corr(dti.tract = dat_harmonized$CCs_rd.combat, my.ylab="CCs RD",less=FALSE,my.ylim = c(0.8,1.4))

ilf = my.function.ga.corr(dti.tract = dat_harmonized$ILF_rd.combat, my.ylab="ILF RD",less=FALSE,my.ylim = c(0.8,1.4))

ifof = my.function.ga.corr(dti.tract = dat_harmonized$IFOF_rd.combat, my.ylab="IFOF RD",less=FALSE,my.ylim = c(0.8,1.4))

af = my.function.ga.corr(dti.tract = dat_harmonized$AF_rd.combat, my.ylab = "AF RD",my.ylim = c(0.8,1.4),less=FALSE)

uf = my.function.ga.corr(dti.tract = dat_harmonized$UF_rd.combat, my.ylab="UF RD",less=FALSE,my.ylim = c(0.8,1.4))

atr = my.function.ga.corr(dti.tract = dat_harmonized$ATR_rd.combat, my.ylab="ATR RD",less=FALSE,my.ylim = c(0.8,1.4))

pt = my.function.ga.corr(dti.tract = dat_harmonized$PT_rd.combat, my.ylab="PT RD",print.legend = FALSE,less=FALSE,estMinCorrAge=FALSE,estMaxCorrAge=FALSE,my.ylim = c(0.8,1.4))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()
```






```{r}

tiff('../Figures/TractSpecific_GestCorrectedAgeRD.tiff',width=1080,height=1280,pointsize=18)
par(mfrow=c(4,3))

#pdf(file='../Figures/TractSpecific_GestCorrectedAgeRD.pdf')

fx = my.function.ga.corr(dti.tract = dat_harmonized$Fx_rd.combat, my.ylab = "FX RD",less=FALSE,my.ylim = c(0.8,1.4),estMinCorrAge = FALSE,estMaxCorrAge=FALSE)

ci = my.function.ga.corr(dti.tract = dat_harmonized$Ci_rd.combat, my.ylab = "Ci RD",my.ylim = c(0.8,1.4),less=FALSE)

ccg = my.function.ga.corr(dti.tract = dat_harmonized$CCg_rd.combat, my.ylab = "CCg RD",my.ylim = c(0.8,1.4),less=FALSE)

ccb = my.function.ga.corr(dti.tract = dat_harmonized$CCb_rd.combat, my.ylab="CCb RD",my.ylim = c(0.8,1.4),less=FALSE)

ccs = my.function.ga.corr(dti.tract = dat_harmonized$CCs_rd.combat, my.ylab="CCs RD",less=FALSE,my.ylim = c(0.8,1.4))

ilf = my.function.ga.corr(dti.tract = dat_harmonized$ILF_rd.combat, my.ylab="ILF RD",less=FALSE,my.ylim = c(0.8,1.4))

ifof = my.function.ga.corr(dti.tract = dat_harmonized$IFOF_rd.combat, my.ylab="IFOF RD",less=FALSE,my.ylim = c(0.8,1.4))

af = my.function.ga.corr(dti.tract = dat_harmonized$AF_rd.combat, my.ylab = "AF RD",my.ylim = c(0.8,1.4),less=FALSE)

uf = my.function.ga.corr(dti.tract = dat_harmonized$UF_rd.combat, my.ylab="UF RD",less=FALSE,my.ylim = c(0.8,1.4))

#atr = my.function.ga.corr(dti.tract = dat_harmonized$ATR_rd.combat, my.ylab="ATR RD",less=FALSE,my.ylim = c(0.8,1.4))

atr = my.function.ga.corr(dti.tract = dat_harmonized$ATR_rd.combat, my.ylab="ATR RD",less=FALSE,my.ylim = c(0.8,1.4),estMinCorrAge = FALSE,estMaxCorrAge=FALSE)
# hack with RD:
abline(v=199,col=cColors[3],lwd=2,lty=2)

pt = my.function.ga.corr(dti.tract = dat_harmonized$PT_rd.combat, my.ylab="PT RD",print.legend = FALSE,less=FALSE,my.ylim = c(0.8,1.4))

plot.new()
  cColors = viridis::plasma(15)
  
legend(x='topleft',legend=c('GA=35','GA=36','GA=37','GA=38','GA=39','GA=40','GA=41'),lty=1,lwd=2,col=cColors[seq(1,13,by=2)],bty='n',cex=1.2)

dev.off()

```



# Additional inspection of models

Check the simultaneous confidenc bands for MD ATR, where the interaction GA*Corrected age was p=0.003:
```{r}
model.atr.md = gam(ATR_md.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
plot_diff(model.atr.md,view='Corr_age',comp=list(GA=c(35,40)),sim.ci = TRUE,xlab='Corrected age (days)')
```
Conclusion: Gestational age has little effect. In ATR, the approximate pvalue is significant, but the difference plot shows little effect. The estimated effect is a bit hard to interpret: at age corrected 0 days, GA=35 are more developed (lower MD), but by 200 days, they are less developed. Note: also checked without sex, and result did not change. 


In MD PT, main effect of GA has p=0.05 in tensor spline model. 
```{r}
summary(gam(PT_md.combat~s(Corr_age,k=9)+s(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized))
```

Check if GA is significant when there is not bivariate spline:
```{r}
model.pt.md = gam(PT_md.combat~s(Corr_age,k=9)+s(GA,k=5)+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
plot_diff(model.pt.md,view='Corr_age',comp=list(GA=c(35,40)),sim.ci = TRUE,xlab='Corrected age (days)')
summary(model.pt.md)
# step into function to examine model diagnostics:
gam.check(model.pt.md)
```
Simultaneous confidence intervals broadly overlapping. The approximate p value of GA is smaller (p=0.02), but I defer to the simultaneous confidence bands. 

Investigate the small region of significant for ATR in RD:
```{r}
model.atr.rd = gam(ATR_rd.combat~ti(Corr_age,k=9)+ti(GA,k=5)+ti(Corr_age,GA,k=c(9,5))+Sex+s(Unique_ID,bs='re'),method='REML',data=dat_harmonized)
plot_diff(model.atr.rd,view='Corr_age',comp=list(GA=c(35,40)),sim.ci = TRUE,xlab='Corrected age (days)')
plot_diff(model.atr.rd,view='Corr_age',comp=list(GA=c(36,40)),sim.ci = TRUE,xlab='Corrected age (days)')

```