---
title: "White Matter Development Data Preparation and Harmonization"
author: "Benjamin Risk"
# date: "2/14/2022"
date: 12/16/2025
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data prep



```{r,results=FALSE}
library(RNifti)
library(readxl)
library(dplyr)
library(stringr)

a = proc.time()
atlas = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/Register_Tracts_from_Template_to_Current_Space/Template_and_Tracts_Original_Asymmetry/tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz')
image(atlas[,,69])
```

Update this atlas with the updated arcuate fasciculus that corrects the R-L issues
```{r}
```

overwrite tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz with new AF:

```{r}
af_liberal=TRUE

if(af_liberal) {
  af = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/AF_liberal/AF_L_R_liberal_flipped_Warp.nii.gz')
  
  # Based on visual inspection of AF_L_R_liberal_flipped_Warp.nii.gz, "1" corresponds to AF right, and "2" to AF left. Correct the labels to correspond to tracts_all_atlas_asymmetry_xyflipped_2_TnP2020.nii.gz:
  
  table(af)
  af[af==2]=9
  af[af==1]=2
  af[af==9]=1
  atlas_old=atlas
  atlas[atlas_old==1 | atlas_old==2] = 0
  atlas[af==1] = af[af==1]
  atlas[af==2] = af[af==2]
}
table(atlas)

```

read in example participant:
```{r}
sub_fa = readNifti('~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/00008-03-02_CSI20151218_tensorup_s_m_aff_diffeo_masked_fa.nii.gz')
```
ben's notes: I checked, and the files  00008-03-02_CSI20151218_tensorup_s_m_aff_diffeo_masked_fa.nii represents the not harmonized data. 

read in labels and "symmetric" data used in main analysis for auditing:
```{r}
labels = read_xlsx("~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/Table_Included_tracts reorderedCol.xlsx")
labels = labels%>%rename(ROI_number = `Intensity Index (Asymmetry)`,ROI_name=`Abbreviations`)
# sort by ROI_number:
labels = labels[order(labels$ROI_number),]

dat_harmonized = read.csv('dat_harmonized_old.csv')
actual_ids = dat_harmonized[!duplicated(dat_harmonized$Actual_ID),c('Actual_ID')]
metrics = c('fa','tr','ad','rd')
var_names = labels$ROI_name
roi_labels = labels$ROI_number
```

Compute asymmetric data:
  
```{r}  
# A function that returns the file path for a given subject ID and metric.
file_template <- "~/Dropbox/WhiteMatterDevelopment/nifti/diffeo_warped/harmonized_dti_data/%s_tensorup_s_m_aff_diffeo_masked_%s.nii.gz"                                
# Example expected paths like "/data/dti/ACT123/fa.nii.gz"
file_for <- function(id, metric) sprintf(file_template, id, metric)
# making nice progress here, working through chat gpt code

stopifnot(length(var_names) == length(roi_labels))
stopifnot(is.array(atlas))

# Precompute voxel indices for each ROI (fast later)
roi_index_list <- lapply(roi_labels, function(lbl) which(atlas == lbl))
names(roi_index_list) <- var_names


# Helper safe mean
.safe_mean <- function(x) if (length(x)) mean(x, na.rm = TRUE) else NA_real_
# ben note: see if we can add a tolerance here


# ---- Prepare output frame: Actual_ID + 19*4 = 76 metric columns ----
all_metric_cols <- as.vector(outer(var_names, metrics, function(n, m) paste0(n, "_", m)))
out <- data.frame(Actual_ID = actual_ids, matrix(NA_real_, nrow = length(actual_ids), ncol = length(all_metric_cols)))
colnames(out)[-1] <- all_metric_cols

# ---- Core loop: actual_ids × metrics ----
for (i in seq_along(actual_ids)) {
  id <- actual_ids[i]

  # Optional: message progress
  if (i %% 10 == 0) message("Scan ", i, "/", length(actual_ids), ": ", id)

  for (met in metrics) {
    img_path <- file_for(id, met)
    if (!file.exists(img_path)) {
      warning("Missing file: ", img_path, " (leaving NAs for this subject/metric)")
      next
    }

    img <- readNifti(img_path)

    # Dimension check (basic guard)
    if (!all(dim(img) == dim(atlas))) {
      stop("Dimension mismatch for subject ", id, ", metric ", met,
           ": img dims ", paste(dim(img), collapse = "x"),
           " vs atlas dims ", paste(dim(atlas), collapse = "x"))
    }

    # Compute means per ROI using precomputed indices
    roi_means <- vapply(roi_index_list, function(idx) .safe_mean(img[idx]), numeric(1))
    # to remove 0s: .safe_mean(img[idx][img[idx] != 0])
    
    # Place into correct columns
    cols_this_metric <- paste0(names(roi_means), "_", met)
    out[i, cols_this_metric] <- roi_means
    }
}

out <- out %>%
  rename_with(.fn = ~ sub("_tr$", "_md", .x), .cols = ends_with("_tr")) %>%
  mutate(across(ends_with("_md"), ~ .x / 3))

# excellent, i checked corpus collosum values for the first three scans and they match dat_harmonized!
proc.time()-a
```


## Calculate the symmetric (non-lateralized) version for AF, ATR, Ci, Fx, IFOF, ILF, PT, and UF:

```{r}
# ---- voxel counts for each ROI (fixed across subjects) ----
roi_nvox <- vapply(roi_index_list, length, integer(1))  # names match var_names
metrics=c('fa','md','ad','rd')

# ---- identify tracts with L/R versions based on var_names ----
bases_L <- str_match(var_names, "^(.*)_L$")[,2]
bases_R <- str_match(var_names, "^(.*)_R$")[,2]
lr_bases <- intersect(na.omit(bases_L), na.omit(bases_R))

# Safety: show what will be combined
message("Combining L/R for: ", paste(lr_bases, collapse = ", "))

# ---- add weighted-mean combined columns for each metric ----
for (b in lr_bases) {
  nL <- roi_nvox[paste0(b, "_L")]
  nR <- roi_nvox[paste0(b, "_R")]
  denom <- nL + nR

  for (met in metrics) {
    colL <- paste0(b, "_L_", met)
    colR <- paste0(b, "_R_", met)
    colC <- paste0(b, "_",   met)
    out[[colC]] <- (out[[colL]] * nL + out[[colR]] * nR) / denom
  }
}
```

Calculate whole brain DTI:
```{r}
# ---- Whole-brain weighted means (STRICT: any NA => NA) ----

for (met in metrics) {
  roi_cols <- paste0(var_names, "_", met)   # e.g., AF_L_fa, ..., UF_R_fa
  w <- roi_nvox[var_names]                  # voxel counts aligned to var_names
  W <- sum(w)

  X <- as.matrix(out[, roi_cols, drop = FALSE])

  # If any ROI is NA in a row, WB is NA for that subject
  out[[paste0("WB_", met)]] <-
    as.numeric(X %*% w) / W

  
  stopifnot(all(colnames(out)[match(paste0(var_names, "_", metrics[1]),
                                  colnames(out))] %in% colnames(out)))
  stopifnot(length(w) == ncol(X))
}
```

Audit: check how new whole brain measures compare to old, and check for a couple ROIs:

```{r}
plot(dat_harmonized$WB_fa~out$WB_fa)
summary(lm(dat_harmonized$WB_fa~out$WB_fa))

plot(dat_harmonized$WB_md~out$WB_md)
plot(dat_harmonized$WB_ad~out$WB_ad)
plot(dat_harmonized$WB_rd~out$WB_rd)

plot(dat_harmonized$IFOF_fa~out$IFOF_fa)

plot(dat_harmonized$UF_ad~out$UF_ad)

plot(dat_harmonized$CCb_md~out$CCb_md)
```
Processing reproduces Longchuan's calculations for ROIs that should match. WB very similar but slightly different due to modified AF.

We pass the audit. Active (not archive) version of this code does not contain merge to old dat_harmonized_old



