---
title: "White Matter Development Data Preparation and Harmonization"
author: "Benjamin Risk"
# date: "2/14/2022"
date: 07/31/2025
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Updates:
-Added Family ID
-Added Actual_ID which is used in merge to motion data
-Added mean absolute and relative motion

## Data prep

First, stack the whole brain and the ROI data. Also convert trace (tr) to mean diffusivity (md) by dividing the tr by 3. 

```{r,results=FALSE}

# NOTE: deleted a space and copy and paste to get columns to align
dat.a = read.delim('../Data/dti_data_extracted_by_tracts_all_atlas_symmetry_xyflipped_2_TnP2020.txt',sep=" ",stringsAsFactors = TRUE)

dat.b = read.delim('../Data/dti_data_extracted_by_tracts_all_atlas_symmetry_xyflipped_2_TnP2020_bin.txt',sep=" ",stringsAsFactors = TRUE)

all(names(dat.a)==names(dat.b))

dat = rbind(dat.a,dat.b)

# replace trace with MD:
dat$DTI_type = as.character(dat$DTI_type)
dat$DTI_val[dat$DTI_type=='tr'] = dat$DTI_val[dat$DTI_type=='tr']/3
dat$DTI_type[dat$DTI_type=='tr'] = 'md'
head(dat$DTI_val[dat$DTI_type=='md'])
dat$DTI_type = as.factor(dat$DTI_type)
dat$Scanner = as.factor(ifelse(dat$ifPrisma==0,'trio','prisma'))

table(dat$ROI)
table(dat$ROI,dat$DTI_type)
table(dat[dat$DTI_type=='fa' & dat$ROI=='CCb','Unique_ID'])
# participants vary from 1 to 3 visits
table(table(dat[dat$DTI_type=='fa' & dat$ROI=='CCb','Unique_ID']))
# 41 infants have 1 visit, 26 have 2, and 12 have 3
```


## Added 14 January 2024:

There are issues with Corr_age for one infant. Replace with correct corrected age.

```{r}
# evidence that correction is needed:
dat_short = dat[,c('Unique_ID','chor_age','Corr_age','GA')]
dat_short$my_correctedage = dat$chor_age-(40-dat$GA)*7
dat_short$diff = dat_short$Corr_age - dat_short$my_correctedage

dat_short[which(abs(dat_short$diff)>1),]
# note: GA=37.3 for one infant is correct

# overwrite the old Corr_age with correct Corr_age:
dat$Corr_age = dat$chor_age-(40-dat$GA)*7
```


## Make data wide

Create a dataset with each ROI by DTI metric as a column. The data are provided in long format, and this converts the data to wide format. This also stacks that whole brain data with the regional data. 

```{r}
library(tidyr)
names(dat)
dat$feature_names = paste0(dat$ROI,'_',dat$DTI_type)
dat2 = dat[,c('Actual_ID','DOB','GA','chor_age','Corr_age','Sex','Unique_ID','DTI_val','ifPrisma','Scanner','feature_names')]
dat_wide = dat2%>%pivot_wider(names_from=feature_names,values_from='DTI_val')
```

## Apply longitudinal combat and investigate scanner effects

Construct a b-spline for corrected age. Then longitudinal combat will use this spline for corrected age. We used df=5 to capture a relatively simple curve, since the effective degrees of freedom from initial GAMMs indicated effective degrees of freedom of 3-4 in the penalized spline approach. Based on inspection, the results appear relatively robust to this choice, although the spline may help to reduce some scanner effects. 
Note, longitudinal combat models the batch effect on the mean as an intercept, i.e., shift up or down in the developmental trajectory. Current harmonization methods do not allow batch effects to cause different curves. Thus, it is important to check that there are not substantial interactions between batch (here, scanner) and the develomental trajectory, i.e., that the curves do not substantially differ on the trio and prisma scanners.

```{r}
library(longCombat)
# reshape wide to get input features for longitudinal combat
feature_names=names(dat_wide)[10:length(dat_wide)]
detach('package:tidyr',unload=TRUE)
degreesfreedom=5
temp = splines::bs(dat_wide$Corr_age,df=degreesfreedom)
temp = data.frame(temp)
names(temp) = paste0("Corr_age",1:degreesfreedom)
dat_wide = cbind(dat_wide,temp)
  
  # check the spline fit:
  # model.test = lm(dat_wide$AF_fa~as.matrix(temp))
  # plot(predict(model.test)~dat_wide$Corr_age)
     
dat_combat <- longCombat(idvar='Unique_ID', 
                             timevar='Corr_age',
                             batchvar='Scanner', 
                             features=feature_names,
                             formula='Corr_age1 + Corr_age2+ Corr_age3 + Corr_age4 + Corr_age5+Sex', 
                            #formula='Corr_age + Sex', # note: linear vs spline seem similar
                             ranef='(1|Unique_ID)',
                             verbose=FALSE,
                             data=dat_wide)
dat_harmonized <- merge(dat_wide, dat_combat$data_combat, by=c('Unique_ID', 'Corr_age','Scanner'))

```

## Added 7/30/2025:

Create family ID variable:

```{r}
# how many pairs of twins due we have:
dat_harmonized$Family_ID = factor(substr(dat_harmonized$Unique_ID,1,5))

library(dplyr)
dat_harmonized <- dat_harmonized %>%
  distinct(Unique_ID, .keep_all = TRUE) %>%       # Keep one row per Unique_ID
  count(Family_ID, name = "n_in_family") %>%       # Count number of Unique_IDs per FamilyID
  right_join(dat_harmonized, by = "Family_ID") %>% # Join back to full data
  mutate(Sibling = ifelse(n_in_family > 1, 1, 0))     # Twin = 1 if FamilyID occurs >1

table(dat_harmonized$n_in_family)
# we have 9 pairs of siblings
# we have 30 scans from twins

dat_harmonized <- dat_harmonized %>%
  distinct(Unique_ID, .keep_all = TRUE) %>%
  count(Family_ID, DOB, name = "n_same_family_dob") %>%
  right_join(dat_harmonized, by = c("Family_ID", "DOB")) %>%
  mutate(Twin = if_else(n_same_family_dob > 1, 1, 0))

#View(dat_harmonized%>%filter(Twin==1)%>%select(Family_ID,Unique_ID,DOB,Sibling,Twin)%>%distinct())

```
We have 5 pairs of twins.

## Added 7/31/25:

Calculate Mean Absolute and Relative Motion:

```{r}
motion_data = dat_harmonized[!duplicated(dat_harmonized$Actual_ID),c('Actual_ID','Unique_ID')]
motion_data$Absolute_RMS=motion_data$Mean_RMS=NA

for (i in 1:nrow(motion_data)) {
  rms=read.table(paste0('/home/ben/Dropbox/WhiteMatterDevelopment/EddyMotionParameters/',motion_data$Actual_ID[i],'/eddy_corrected_dataup.eddy_movement_rms'))
  motion_data[i,'Absolute_RMS'] = mean(rms[2:nrow(rms),1])
  motion_data[i,'Mean_RMS'] = mean(rms[2:nrow(rms),2])
}
motion_data = motion_data%>%select(-Unique_ID)
dat_harmonized=merge(dat_harmonized,motion_data)
```

Write file to disk:
```{r}
write.csv(dat_harmonized,file='dat_harmonized_old.csv',row.names=FALSE)
```